---
title: 'Metabolite signature - dementia biomarker - clinical diagnostic mediation evaluation'
author: 
    - name: 'Marc Clos-Garcia'
      affiliation: 'Steno Diabetes Center Copenhagen, Denmark'
      email: 'marc.clos.garcia@regiong.dk'
date: last-modified
date-format: "D MMMM, YYYY"
format: 
    html:
        toc: true
        toc-depth: 3
        number-sections: true
        code-fold: true
        code-summary: 'Click to expand/collapse code'
        theme: cosmo
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

In this report, we aimed to disentangle the relationships between the set of stable metabolites that the ElasticNet modelling, coupled to bootstrapping, identified as *predictive* of tTau, pTau and Aβ42. To do so, we have used Structural Equation Modelling. This framework was chosen over traditional mediation for several reasons. First, it allowed for the construction of latent variables, aggregating the signal from multiple highly correlated metabolites (identified via Elastic Net) into a single biological construct. This minimizes individual measurement error and addresses the high dimensionality of the data. Second, while the initial metabolite selection was informed by association with biomarkers, the SEM framework provides a rigorous test of the directional structural path, mitigating concerns of data circularity by requiring the data to conform to a specific mediated architecture. Finally, to ensure the validity of our findings given the limited sample sizes of certain clinical subgroups, we utilized non-parametric bootstrapping (1,000 iterations) to generate 95% confidence intervals and p-values for the indirect effects, bypassing the normality assumptions required by traditional mediation tests.

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
# import libraries
library(here)
library(tidyverse)
library(ggplot2)
library(reticulate)
```

```{r directories}
getwd()
models_dir = here('results/modelling/')
```

```{r utilities}
feature_labels = data.frame(
  "current_names" = c('d.Glucose', 'gluc.erythitol', 'd.Galactose', 'gluc.sorbitol', 'D.....Xylose', 'D.....Ribofuranose',
                      'Sorbitol', 'L.....Arabitol', 'D.Threitol', 'meso.Erythritol', 'X3.Deoxy.erythro.pentitol',
                      'X1.5.Anhydrohexitol', 'Ribonic.acid', 'L.Threonic.acid', 'Glyceric.acid'),
  "plotting_names" = c('d-Glucose', 'd-Glucose / Erythritol', 'd-Galactose', 'd-Glucose / Sorbitol', 'D-(-)-Xylose',
                       'D-Ribofuranose', 'Sorbitol', 'L-(-)-Arabitol', 'D-Threitol', 'meso-Erythritol', 
                       '3-Deoxy-erythro-pentitol', '1,5-Anhydrohexitol', 'Ribonic acid', 'L-Threonic acid', 'Glyceric acid')
)
```


```{r data load}
# import data
metadata = read.table(here('data/csf_ratios2.csv'), sep = ";", header = T)
metadata = metadata |> dplyr::select('DDBBno', 'Gender', 'AgeAtVisit', 'fullClass', 'TAU', 'PTAU', 'ABETA42')

## encode gender
metadata <- metadata |> dplyr::mutate(Gender_encoded = ifelse(Gender == "Male", 1, 0))

## transform & scale tTau, pTay and AB42 as we did before for consistency
metadata$TAU <- scale(log(metadata$TAU))
metadata$PTAU <- scale(log(metadata$PTAU))
metadata$ABETA42 <- scale(metadata$ABETA42)

head(metadata)
```

# Mediation results 

::: {.panel-tabset}

## tTau mediation

### Mediation results
```{r mediation table}
# read in all tTau results files
files <- list.files(models_dir, pattern = "SEM_mediation_table_.*_TAU\\.csv", full.names = TRUE)

# Concatenate all
df_list <- lapply(files, read.csv)
master_df <- bind_rows(df_list)

# Define custom order
contrast_order <- c("HC_vs_MCI", "HC_vs_AD", "HC_vs_VaD", "MCI_vs_AD", "MCI_vs_VaD", "AD_vs_VaD")
master_df$Contrast <- factor(master_df$Contrast, levels = contrast_order, ordered = TRUE)

# Format Estimates and CIs
format_result <- function(estimate, ci_lower, ci_upper) {
  if (is.na(ci_lower)) {  # For Proportion/Ratio rows
    return(sprintf("%.3f", estimate))
  }
  return(sprintf("%.3f [%.3f, %.3f]", estimate, ci_lower, ci_upper))
}

master_df$`Estimate (95% CI)` <- mapply(format_result, 
                                         master_df$Estimate, 
                                         master_df$CI_Lower, 
                                         master_df$CI_Upper)

# Sort by Contrast and Effect
table_display <- master_df |> arrange(Contrast, Effect)

# Rename the 'Effect' values for better readability
table_display$Effect <- recode(table_display$Effect,
                                "ind" = "Indirect (via Tau)",
                                "dir" = "Direct Effect",
                                "tot" = "Total Effect")

# Display with Grouping
final_table <- table_display |> 
  dplyr::select(Contrast, Effect, `Estimate (95% CI)`, p_val)

colnames(final_table) <- c("Clinical Contrast", "Pathway", "Beta [95% CI]", "p-value")
knitr::kable(final_table)
```

### Metabolite contributions (loadings)

```{r loadings, fig.width = 12, fig.height = 6.5}
# Loadings files
loadings_files <- list.files(models_dir, pattern = "SEM_mediation_loadings_.*_TAU\\.csv", full.names = TRUE)

df_loadings <- list()
for (f in loadings_files) {
  # Read the individual CSV
  temp_df <- read.csv(f)
  
  # Extract the contrast from filename
  contrast_name <- basename(f) |> 
    str_replace("SEM_mediation_loadings_", "") |> 
    str_replace("_TAU\\.csv", "")
  
  # Add the column to the temporary dataframe
  temp_df$Contrast <- contrast_name
  df_loadings[[length(df_loadings) + 1]] <- temp_df
}

# Concatenate all
master_loads_df <- bind_rows(df_loadings)

# Reorder columns to put Contrast first
master_loads_df <- master_loads_df |> 
  dplyr::select(Contrast, everything())
master_loads_df <- master_loads_df |> 
  dplyr::left_join(feature_labels, by = c("Metabolite" = "current_names")) |> 
  dplyr::select(-Metabolite) |> dplyr::select(Metabolite = plotting_names, Contrast, everything())

master_loads_df$Contrast <- factor(master_loads_df$Contrast, levels = contrast_order)

master_loads_df$pval_star <- case_when(
  master_loads_df$fdr_p <= 0.001 ~ "***",
  master_loads_df$fdr_p <= 0.01 ~ "**",
  master_loads_df$fdr_p <= 0.05 ~ "*",
  master_loads_df$fdr_p > 0.05 ~ "ns",
)

ggplot(master_loads_df, aes(Median_Loading, Metabolite)) +
  geom_point(shape = 15) +
  geom_linerange(aes(xmin = CI_Lower, xmax = CI_Upper)) +
  geom_vline(xintercept = 0, linetype = 'dashed') +
  geom_text(aes(label = pval_star), 
            x = 2.5,       # Fixed x position
            hjust = 0,
            size = 3,
            na.rm = TRUE,
            fontface = 'italic') +
  facet_wrap(~Contrast) +
  theme_minimal() +
  labs(title = "Loadings for the tTau SEM models", 
        x = "Median loading contribution")

master_loads_df_ttau <- master_loads_df |> dplyr::filter(Contrast == "HC_vs_AD")
master_loads_df_ttau$biomarker <- "tTau"
```

### ROC curves
```{r rocs, fig.width=22, fig.height=10}
# import python pickle module
pickle <- reticulate::import("pickle")
py_open <- import_builtins()$open

roc_files <- c(file.path(models_dir, "SEM_mediation_roc_HC_vs_MCI_TAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_HC_vs_AD_TAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_HC_vs_VaD_TAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_MCI_vs_AD_TAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_MCI_vs_VaD_TAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_AD_vs_VaD_TAU.pkl"))

roc_curves <- list()
for(f in roc_files){
  contrast <- sub(".*roc_(.*?)_[^_]+\\.pkl$", "\\1", f)
  roc_x <- pickle$load(py_open(f, "rb")) 
  # extract AUCs
  roc_x_auc <- lapply(roc_x, function(r){
    r$auc
  })
  auc_text <- paste(
    "AUC values:",
    sprintf("Biomarker:   %.3f", roc_x_auc$Biomarker),
    sprintf("Metabolites: %.3f", roc_x_auc$Metabolites),
    sprintf("Combined:    %.3f", roc_x_auc$Combined),
    sep = "\n"
  )
  # extract FPRs (x-axis)
  roc_x_fpr <- lapply(roc_x, function(r){
    r$fpr
  })
  # extract TPRs (y-axis)
  roc_x_tpr <- lapply(roc_x, function(r){
    r$tpr
  })
  ROCs <- cbind(
    unlist(roc_x_fpr) |> as.data.frame() |> tibble::rownames_to_column("Curve") |> dplyr::mutate(Curve = gsub("[0-9+]", "", Curve)),
    unlist(roc_x_tpr) |> as.data.frame() |> tibble::rownames_to_column("Curve2") |> dplyr::mutate(Curve2 = gsub("[0-9+]", "", Curve2))
  ) |> 
    dplyr::select(-Curve2)
  
  names(ROCs) <- c("Curve", "FPR", "TPR")
  ROCs$Curve <- factor(ROCs$Curve, levels = c("Biomarker", "Metabolites", "Combined"))
  roc.p <- ggplot(ROCs, aes(FPR, TPR, group = Curve, color = Curve)) +
    geom_line(size = 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_minimal() +
    annotate("label",
             x = 0.7, y = 0.15,      # adjust position as needed
             label = auc_text,
             hjust = 0,
             size = 4) +
    scale_color_manual(values = c("#C0392B",
                                  "#2980B9",
                                  "#27AE60")) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1)) +
    labs(x = "False Positive Rate (1-specificity)",
         y = "True Positive Rate (sensitivity)", 
         title = paste0("ROC curve for the contrast: ", contrast))
  roc_curves[[f]] <- roc.p
}

cowplot::plot_grid(plotlist = roc_curves)

  
```


## pTau mediation

### Mediation results
```{r mediation table ptau}
# read in all PTau results files
files <- list.files(models_dir, pattern = "SEM_mediation_table_.*_PTAU\\.csv", full.names = TRUE)

# Concatenate all
df_list <- lapply(files, read.csv)
master_df <- bind_rows(df_list)

# Define custom order
master_df$Contrast <- factor(master_df$Contrast, levels = contrast_order, ordered = TRUE)

# Format Estimates and CIs
format_result <- function(estimate, ci_lower, ci_upper) {
  if (is.na(ci_lower)) {  # For Proportion/Ratio rows
    return(sprintf("%.3f", estimate))
  }
  return(sprintf("%.3f [%.3f, %.3f]", estimate, ci_lower, ci_upper))
}

master_df$`Estimate (95% CI)` <- mapply(format_result, 
                                         master_df$Estimate, 
                                         master_df$CI_Lower, 
                                         master_df$CI_Upper)

# Sort by Contrast and Effect
table_display <- master_df |> arrange(Contrast, Effect)

# Rename the 'Effect' values for better readability
table_display$Effect <- recode(table_display$Effect,
                                "ind" = "Indirect (via pTau)",
                                "dir" = "Direct Effect",
                                "tot" = "Total Effect")

# Display with Grouping
final_table <- table_display |> 
  dplyr::select(Contrast, Effect, `Estimate (95% CI)`, p_val)

colnames(final_table) <- c("Clinical Contrast", "Pathway", "Beta [95% CI]", "p-value")
knitr::kable(final_table)
```

### Metabolite contributions (loadings)

```{r loadings ptau, fig.width = 12, fig.heigth = 6.5}
# Loadings files
loadings_files <- list.files(models_dir, pattern = "SEM_mediation_loadings_.*_PTAU\\.csv", full.names = TRUE)

df_loadings <- list()
for (f in loadings_files) {
  # Read the individual CSV
  temp_df <- read.csv(f)
  
  # Extract the contrast from filename
  contrast_name <- basename(f) |> 
    str_replace("SEM_mediation_loadings_", "") |> 
    str_replace("_PTAU\\.csv", "")
  
  # Add the column to the temporary dataframe
  temp_df$Contrast <- contrast_name
  df_loadings[[length(df_loadings) + 1]] <- temp_df
}

# Concatenate all
master_loads_df <- bind_rows(df_loadings)

# Reorder columns to put Contrast first
master_loads_df <- master_loads_df |> 
  dplyr::select(Contrast, everything())
master_loads_df <- master_loads_df |> 
  dplyr::left_join(feature_labels, by = c("Metabolite" = "current_names")) |> 
  dplyr::select(-Metabolite) |> dplyr::select(Metabolite = plotting_names, Contrast, everything())

master_loads_df$Contrast <- factor(master_loads_df$Contrast, levels = contrast_order)

master_loads_df$pval_star <- case_when(
  master_loads_df$fdr_p <= 0.001 ~ "***",
  master_loads_df$fdr_p <= 0.01 ~ "**",
  master_loads_df$fdr_p <= 0.05 ~ "*",
  master_loads_df$fdr_p > 0.05 ~ "ns",
)

ggplot(master_loads_df, aes(Median_Loading, Metabolite)) +
  geom_point(shape = 15) +
  geom_linerange(aes(xmin = CI_Lower, xmax = CI_Upper)) +
  geom_vline(xintercept = 0, linetype = 'dashed') +
  geom_text(aes(label = pval_star), 
            x = 2.5,       # Fixed x position
            hjust = 0,
            size = 3,
            na.rm = TRUE,
            fontface = 'italic') +
  facet_wrap(~Contrast) +
  theme_minimal() +
  labs(title = "Loadings for the pTau SEM models", 
        x = "Median loading contribution")

master_loads_df_ptau <- master_loads_df |> dplyr::filter(Contrast == "HC_vs_AD")
master_loads_df_ptau$biomarker <- "pTau"
```

### ROC curves
```{r rocs ptau, fig.width=22, fig.height=10}
# import python pickle module
pickle <- reticulate::import("pickle")
py_open <- import_builtins()$open

roc_files <- c(file.path(models_dir, "SEM_mediation_roc_HC_vs_MCI_PTAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_HC_vs_AD_PTAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_HC_vs_VaD_PTAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_MCI_vs_AD_PTAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_MCI_vs_VaD_PTAU.pkl"),
               file.path(models_dir, "SEM_mediation_roc_AD_vs_VaD_PTAU.pkl"))

roc_curves <- list()
for(f in roc_files){
  contrast <- sub(".*roc_(.*?)_[^_]+\\.pkl$", "\\1", f)
  roc_x <- pickle$load(py_open(f, "rb")) 
  # extract AUCs
  roc_x_auc <- lapply(roc_x, function(r){
    r$auc
  })
  auc_text <- paste(
    "AUC values:",
    sprintf("Biomarker:   %.3f", roc_x_auc$Biomarker),
    sprintf("Metabolites: %.3f", roc_x_auc$Metabolites),
    sprintf("Combined:    %.3f", roc_x_auc$Combined),
    sep = "\n"
  )
  # extract FPRs (x-axis)
  roc_x_fpr <- lapply(roc_x, function(r){
    r$fpr
  })
  # extract TPRs (y-axis)
  roc_x_tpr <- lapply(roc_x, function(r){
    r$tpr
  })
  ROCs <- cbind(
    unlist(roc_x_fpr) |> as.data.frame() |> tibble::rownames_to_column("Curve") |> dplyr::mutate(Curve = gsub("[0-9+]", "", Curve)),
    unlist(roc_x_tpr) |> as.data.frame() |> tibble::rownames_to_column("Curve2") |> dplyr::mutate(Curve2 = gsub("[0-9+]", "", Curve2))
  ) |> 
    dplyr::select(-Curve2)
  
  names(ROCs) <- c("Curve", "FPR", "TPR")
  ROCs$Curve <- factor(ROCs$Curve, levels = c("Biomarker", "Metabolites", "Combined"))
  roc.p <- ggplot(ROCs, aes(FPR, TPR, group = Curve, color = Curve)) +
    geom_line(size = 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_minimal() +
    annotate("label",
             x = 0.7, y = 0.15,      # adjust position as needed
             label = auc_text,
             hjust = 0,
             size = 4) +
    scale_color_manual(values = c("#C0392B",
                                  "#2980B9",
                                  "#27AE60")) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1)) +
    labs(x = "False Positive Rate (1-specificity)",
         y = "True Positive Rate (sensitivity)", 
         title = paste0("ROC curve for the contrast: ", contrast))
  roc_curves[[f]] <- roc.p
}

cowplot::plot_grid(plotlist = roc_curves)

  
```



## Aβ42 mediation

### Mediation results
```{r mediation table ab42}
# read in all Aβ42 results files
files <- list.files(models_dir, pattern = "SEM_mediation_table_.*_ABETA42\\.csv", full.names = TRUE)

# Concatenate all
df_list <- lapply(files, read.csv)
master_df <- bind_rows(df_list)

# Define custom order
master_df$Contrast <- factor(master_df$Contrast, levels = contrast_order, ordered = TRUE)

# Format Estimates and CIs
format_result <- function(estimate, ci_lower, ci_upper) {
  if (is.na(ci_lower)) {  # For Proportion/Ratio rows
    return(sprintf("%.3f", estimate))
  }
  return(sprintf("%.3f [%.3f, %.3f]", estimate, ci_lower, ci_upper))
}

master_df$`Estimate (95% CI)` <- mapply(format_result, 
                                         master_df$Estimate, 
                                         master_df$CI_Lower, 
                                         master_df$CI_Upper)

# Sort by Contrast and Effect
table_display <- master_df |> arrange(Contrast, Effect)

# Rename the 'Effect' values for better readability
table_display$Effect <- recode(table_display$Effect,
                                "ind" = "Indirect (via Aβ42)",
                                "dir" = "Direct Effect",
                                "tot" = "Total Effect")

# Display with Grouping
final_table <- table_display |> 
  dplyr::select(Contrast, Effect, `Estimate (95% CI)`, p_val)

colnames(final_table) <- c("Clinical Contrast", "Pathway", "Beta [95% CI]", "p-value")
knitr::kable(final_table)
```

### Metabolite contributions (loadings)

```{r loadings ab42, fig.width = 12, fig.height = 6.5}
# Loadings files
loadings_files <- list.files(models_dir, pattern = "SEM_mediation_loadings_.*_ABETA42\\.csv", full.names = TRUE)

df_loadings <- list()
for (f in loadings_files) {
  # Read the individual CSV
  temp_df <- read.csv(f)
  
  # Extract the contrast from filename
  contrast_name <- basename(f) |> 
    str_replace("SEM_mediation_loadings_", "") |> 
    str_replace("_ABETA42\\.csv", "")
  
  # Add the column to the temporary dataframe
  temp_df$Contrast <- contrast_name
  df_loadings[[length(df_loadings) + 1]] <- temp_df
}

# Concatenate all
master_loads_df <- bind_rows(df_loadings)

# Reorder columns to put Contrast first
master_loads_df <- master_loads_df |> 
  dplyr::select(Contrast, everything())
master_loads_df <- master_loads_df |> 
  dplyr::left_join(feature_labels, by = c("Metabolite" = "current_names")) |> 
  dplyr::select(-Metabolite) |> dplyr::select(Metabolite = plotting_names, Contrast, everything())

master_loads_df$Contrast <- factor(master_loads_df$Contrast, levels = contrast_order)

master_loads_df$pval_star <- case_when(
  master_loads_df$fdr_p <= 0.001 ~ "***",
  master_loads_df$fdr_p <= 0.01 ~ "**",
  master_loads_df$fdr_p <= 0.05 ~ "*",
  master_loads_df$fdr_p > 0.05 ~ "ns",
)

ggplot(master_loads_df, aes(Median_Loading, Metabolite)) +
  geom_point(shape = 15) +
  geom_linerange(aes(xmin = CI_Lower, xmax = CI_Upper)) +
  geom_vline(xintercept = 0, linetype = 'dashed') +
  geom_text(aes(label = pval_star), 
            x = 2.5,       # Fixed x position
            hjust = 0,
            size = 3,
            na.rm = TRUE,
            fontface = 'italic') +
  facet_wrap(~Contrast) +
  theme_minimal() +
  labs(title = "Loadings for the Aβ42 SEM models", 
        x = "Median loading contribution")

master_loads_df_ab42 <- master_loads_df |> dplyr::filter(Contrast == "HC_vs_AD")
master_loads_df_ab42$biomarker <- "Aβ42"
```

### ROC curves
```{r rocs ABETA42, fig.width=22, fig.height=10}
# import python pickle module
pickle <- reticulate::import("pickle")
py_open <- import_builtins()$open

roc_files <- c(file.path(models_dir, "SEM_mediation_roc_HC_vs_MCI_ABETA42.pkl"),
               file.path(models_dir, "SEM_mediation_roc_HC_vs_AD_ABETA42.pkl"),
               file.path(models_dir, "SEM_mediation_roc_HC_vs_VaD_ABETA42.pkl"),
               file.path(models_dir, "SEM_mediation_roc_MCI_vs_AD_ABETA42.pkl"),
               file.path(models_dir, "SEM_mediation_roc_MCI_vs_VaD_ABETA42.pkl"),
               file.path(models_dir, "SEM_mediation_roc_AD_vs_VaD_ABETA42.pkl"))

roc_curves <- list()
for(f in roc_files){
  contrast <- sub(".*roc_(.*?)_[^_]+\\.pkl$", "\\1", f)
  roc_x <- pickle$load(py_open(f, "rb")) 
  # extract AUCs
  roc_x_auc <- lapply(roc_x, function(r){
    r$auc
  })
  auc_text <- paste(
    "AUC values:",
    sprintf("Biomarker:   %.3f", roc_x_auc$Biomarker),
    sprintf("Metabolites: %.3f", roc_x_auc$Metabolites),
    sprintf("Combined:    %.3f", roc_x_auc$Combined),
    sep = "\n"
  )
  # extract FPRs (x-axis)
  roc_x_fpr <- lapply(roc_x, function(r){
    r$fpr
  })
  # extract TPRs (y-axis)
  roc_x_tpr <- lapply(roc_x, function(r){
    r$tpr
  })
  ROCs <- cbind(
    unlist(roc_x_fpr) |> as.data.frame() |> tibble::rownames_to_column("Curve") |> dplyr::mutate(Curve = gsub("[0-9+]", "", Curve)),
    unlist(roc_x_tpr) |> as.data.frame() |> tibble::rownames_to_column("Curve2") |> dplyr::mutate(Curve2 = gsub("[0-9+]", "", Curve2))
  ) |> 
    dplyr::select(-Curve2)
  
  names(ROCs) <- c("Curve", "FPR", "TPR")
  ROCs$Curve <- factor(ROCs$Curve, levels = c("Biomarker", "Metabolites", "Combined"))
  roc.p <- ggplot(ROCs, aes(FPR, TPR, group = Curve, color = Curve)) +
    geom_line(size = 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    theme_minimal() +
    annotate("label",
             x = 0.7, y = 0.15,      # adjust position as needed
             label = auc_text,
             hjust = 0,
             size = 4) +
    scale_color_manual(values = c("#C0392B",
                                  "#2980B9",
                                  "#27AE60")) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1)) +
    labs(x = "False Positive Rate (1-specificity)",
         y = "True Positive Rate (sensitivity)", 
         title = paste0("ROC curve for the contrast: ", contrast))
  roc_curves[[f]] <- roc.p
}

cowplot::plot_grid(plotlist = roc_curves)

  
```




:::

```{r fig.width=13, fig.height=4.5}
cowplot::plot_grid(
  ggplot(master_loads_df_ttau, aes(Median_Loading, Metabolite)) +
    geom_point(shape = 15) +
    geom_linerange(aes(xmin = CI_Lower, xmax = CI_Upper)) +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    geom_text(aes(label = pval_star, x = CI_Upper), 
              # x = 2.5,       # Fixed x position
              hjust = -0.3,
              size = 3,
              na.rm = TRUE,
              fontface = 'italic') +
    # facet_grid(~biomarker, scales = "free") +
    theme_minimal() +
    labs(
         x = "Median loading contribution"),
  ggplot(master_loads_df_ptau, aes(Median_Loading, Metabolite)) +
    geom_point(shape = 15) +
    geom_linerange(aes(xmin = CI_Lower, xmax = CI_Upper)) +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    geom_text(aes(label = pval_star, x = CI_Upper), 
              # x = 2.5,       # Fixed x position
              hjust = -0.3,
              size = 3,
              na.rm = TRUE,
              fontface = 'italic') +
    # facet_grid(~biomarker, scales = "free") +
    theme_minimal() +
    labs(
         x = "Median loading contribution"),
  ggplot(master_loads_df_ab42, aes(Median_Loading, Metabolite)) +
    geom_point(shape = 15) +
    geom_linerange(aes(xmin = CI_Lower, xmax = CI_Upper)) +
    geom_vline(xintercept = 0, linetype = 'dashed') +
    geom_text(aes(label = pval_star, x = CI_Upper), 
              # x = 2.5,       # Fixed x position
              hjust = -0.3,
              size = 3,
              na.rm = TRUE,
              fontface = 'italic') +
    # facet_grid(~biomarker, scales = "free") +
    theme_minimal() +
    labs(
         x = "Median loading contribution"),
  ncol = 3, labels = c("tTau", "pTau", "Aβ42")
)

```

