---
title: "Lundbeck project - Targeted CSF CGMS metabolomics - Rigshospitalet - QC"
author: "Marc Clos-Garcia"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: united
date: "`r Sys.Date()`"
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
## libraries
library(here)
library(tidyverse)
library(ggpubr)
library(factoextra)
library(VIM)
library(PCAtools)
library(ropls)
library(DT)
library(gridExtra)
library(MASS)
library(ggfortify)
library(DescTools)

## directories
dir_data <- here("data/CSF")
if (!dir.exists(dir_data)) dir.create(dir_data)
dir_res <- here("results/CSF")
if (!dir.exists(dir_res)) dir.create(dir_res)
dir_code <- here("code")
if (!dir.exists(dir_code)) dir.create(dir_code)
```

```{r load and format datasets, echo=FALSE, message=FALSE, warning=FALSE}
#### GC-MS
targeted <- read.table(file.path(dir_data, "0112 LundbeckCSF targeted results.csv"), header = T, sep = "\t", dec = ".")
targeted <- targeted[, -c(1:2)]
names(targeted)[1:5] <- as.character(targeted[1, 1:5])
names(targeted) <- gsub("\\.Results$", "", names(targeted))
targeted <- targeted[-1, ]

##### keep only the final concentration columns (columns with metabolite names)
targeted <- targeted[, !grepl("^X\\.\\d+$", names(targeted))]

###### extract sample numbers
targeted$Name <- gsub("_backup", "", targeted$Name)
targeted$`Data File` <- gsub("_backup", "", targeted$`Data File`)
targeted <- targeted %>%
    tidyr::separate("Data File", sep = "_", into = c("P_number", "P_name", "Date", "Run", "sample", "vial", "injection")) %>%
    dplyr::mutate(sample = ifelse(is.na(sample), Name, sample)) %>%
    dplyr::select(-P_number, -P_name, -Date, -vial, -injection) %>%
    as.data.frame()

#### clinical - sample IDs
mtdt <- read.delim(file.path(dir_data, "Sample codenumbers plus data 2023_01_03_KH.csv"), header = T, sep = ",")

#### IDs matching file
matching <- read.table(file.path(dir_data, "0112_Lundbeck_CSF_QQQ_STENO_ID_DDBBno_each_sample.csv"), header = T, sep = ",", skip = 1)
matching <- matching[, -3]
```


```{r add DDBBno ids to metadata and merge all files, echo=FALSE, message=FALSE, warning=FALSE}
#### add ID to metadata
mtdt <- mtdt %>%
    dplyr::left_join(matching, by = "DDBBno") %>%
    dplyr::select(STENO.ID, DDBBno, everything())
mtdt$STENO.ID <- as.character(mtdt$STENO.ID)

#### add extra metadata
extra_metadata <- readxl::read_excel(file.path(dir_data, "RH_Clinical info BMI_Statin_HRT_2025_06_23.xlsx"), sheet = 1, col_names = T)
extra_metadata <- extra_metadata %>% dplyr::select(DDBBno, BMI, Statins, "Hormone replacement therapy", `Neuropsychology_ACE_Total score`, `Genetics_Risk factors_APOE`, CSF_glucose = "Cerebrospinal fluids_Glucose#mmol/l")

mtdt <- mtdt %>% 
  dplyr::left_join(extra_metadata, by = "DDBBno") %>% 
  dplyr::mutate(Statins = ifelse(Statins == "Nej", 0, 1)) %>% 
  dplyr::mutate("Hormone replacement therapy" = ifelse(`Hormone replacement therapy` == "Nej", 0, 1))

#### merge metabolomics with metadata
targeted <- targeted %>%
    dplyr::left_join(mtdt, by = c("sample" = "STENO.ID")) %>%
    dplyr::select(
        Name, sample, DDBBno, Run, Type, Level, "Acq. Date-Time", Gender, AgeAtVisit, CognitiveSyndrome,
        Diagnosis, Class, MMSE_Score, Cerebrospinal.fluids_Amyloid.beta.42.pg.ml,
        Cerebrospinal.fluids_Phospho.Tau.pg.ml, Cerebrospinal.fluids_Total.Tau.pg.ml, BMI, 
        Statins, "Hormone replacement therapy", "Neuropsychology_ACE_Total score", 
        "Genetics_Risk factors_APOE", CSF_glucose, everything()
    )

## clean up
rm(matching)
```


```{r data transformations, echo=FALSE, message=FALSE, warning=FALSE}
## data transformations
### transform to numbers
targeted[, 23:ncol(targeted)] <- sapply(targeted[, 23:ncol(targeted)], as.numeric) ## convert to numeric
```

# Context
In this report, we perform a quality-control evaluation of the targeted metabolomics dataset performed in CSF. In total, the dataset included up to `r ncol(targeted[, 23:ncol(targeted)])` annotated targeted metabolites. The process presented here includes:

* Evaluating the ratios between blanks and samples
* Evaluation of Coefficient of Variation (CV) in the sample pools
* Identification and processing of duplicate samples
* Missing values evaluation and imputation
* Data normalization
* Identification and removal of outliers

# Check calibration curves
```{r check calbration curves, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=8.75}
targeted_cal <- targeted %>% dplyr::filter(Type == "Cal")
targeted_cal$Name <- factor(targeted_cal$Name, levels = targeted_cal$Name)
targeted_cal <- targeted_cal %>% dplyr::select(-c("Run", "Type", "Level", "Acq. Date-Time", "sample", "DDBBno", "Gender", "AgeAtVisit",
                                                  "CognitiveSyndrome", "Diagnosis", "Class", "MMSE_Score", "Cerebrospinal.fluids_Amyloid.beta.42.pg.ml",
                                                  "Cerebrospinal.fluids_Phospho.Tau.pg.ml", "Cerebrospinal.fluids_Total.Tau.pg.ml", "BMI", "Statins",
                                                  "Hormone replacement therapy", "Neuropsychology_ACE_Total score", "Genetics_Risk factors_APOE",
                                                  "CSF_glucose")) 
targeted_cal %>% 
  reshape::melt() %>%
  ggplot(aes(Name, value)) +
    geom_point() +
    geom_line() +
    facet_wrap(~variable, scales = "free") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))

## remove calibration samples
targeted <- targeted |> dplyr::filter(!grepl("Cal", sample))

```


# Check for duplicates {.tabset}
```{r duplicates, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# sort(table(targeted$sample), decreasing = TRUE) ## we check duplicates by sample name, no duplicates are observed except for blank17, CP17 and RPO17

# Check missing patterns to see if any of them has more missing data
# This is also a type of QCing the data to check how well duplicates correlate
dups.df <- targeted[duplicated(targeted$sample), ]
datatable(dups.df)

dup.samples <- targeted[duplicated(targeted$sample), "sample"]

for (x in dup.samples) {
    cat("## ", x, "\n\n")
    nas <- data.frame(cbind(
        summary(rowSums(is.na(targeted[which(targeted$sample == x)[1], 23:ncol(targeted)]))),
        summary(rowSums(is.na(targeted[which(targeted$sample == x)[2], 23:ncol(targeted)])))
    ))
    names(nas) <- which(targeted$sample == x)
    print(htmltools::tagList(DT::datatable(nas,
        escape = T, height = "100%", class = "cell-border stripe",
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
    cat("\n\n")
    # Plot the two pairs of data
    plot(t(targeted[which(targeted$sample == x), 23:ncol(targeted)]))
    cat("\n\n")
}

## samples look very similar, we'll remove one duplicate randomly
targeted <- targeted[!duplicated(targeted$sample), ]

# clean up
rm(dup.samples, nas)
```


# {.unnumbered .unlisted}
We see `r nrow(dups.df)` duplicates. After evaluating the metabolites distribution across these duplicates, we decide to randomly remove one duplicate for each pair.

# Missing values
```{r NAs, echo=FALSE, message=FALSE, warning=FALSE}
# Check participants with all met data missing
targeted$allNA <- apply(targeted[, 23:ncol(targeted)],
    MARGIN = 1, FUN = function(x) all(is.na(x))
)

summary(targeted$allNA)

# NONE missing- remove allNA column
targeted$allNA <- NULL


# check missingess per individual
summary(rowSums(is.na(targeted[, 23:ncol(targeted)])))
hist(summary(rowSums(is.na(targeted[, 23:ncol(targeted)]))), breaks = 10)
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 0.0000  0.0000  0.0000  0.3636  0.0000  6.0000

row.names(targeted) <- targeted$sample


## Exclude individuals with missing data in more than 20% of metabolites -!!!
targeted.f <- targeted[, 23:ncol(targeted)][which(rowMeans(is.na(targeted[, 23:ncol(targeted)])) <= 0.2), ]

# 0 individuals have more than 20% of missing data-> 251 individuals left

# Exclude metabolites with missing data in more than 20% of metabolites
targeted.f2 <- targeted.f[, which(colMeans(is.na(targeted.f)) <= 0.2), ]

dim(targeted.f2)
# [1]  251 44  left
# no metabolites missing more than 20%

## merge back the filtered dataset
targeted.f <- merge(targeted[, 1:22], targeted.f2, by = "row.names")
targeted.f$Row.names <- NULL
dim(targeted.f)
```

No samples were removed due to the number of missing values (all <20%). 
Missing values were imputed using kNN clustering approach with k=10.

```{r impute missing values, echo=FALSE, message=FALSE, warning=FALSE}
### impute missing values with kNN
targeted.imp <- kNN(targeted.f[, 23:ncol(targeted.f)], k = 10)
targeted.imp <- cbind(targeted.f[, 1:22], targeted.imp)
targeted.imp <- targeted.imp %>%
    dplyr::select(-contains("_imp")) %>%
    as.data.frame()
targeted.imp[, 23:ncol(targeted.imp)] <- sapply(targeted.imp[, 23:ncol(targeted.imp)], as.numeric) # convert metabolite data back to numeric

rm(targeted, targeted.f, targeted.f2)
```


# Data normalization {.tabset}
We normalized the metabolomics data by log10 transforming it. For both the raw and log10 transformed data we have plotted the individual metabolites abundance distribution histograms, pair-by-pair. The PDFs with such histogram plots can be find in `r dir_res`.
We also show, below, the boxplots for the raw and log-transformed data, to rapidly evaluate the effectivness of the data transformation.
```{r Visualize distribution 1, echo=FALSE, message=FALSE, warning=FALSE}
# Vector of all lipids to iterate through
Metabolite_names <- names(targeted.imp)[23:ncol(targeted.imp)]
# Visualize histograms
p <- list()
for (j in Metabolite_names) {
    p.raw <- ggplot(targeted.imp, aes(x = (targeted.imp[, j]))) +
        geom_histogram(aes_string(x = targeted.imp[, j]), bins = 30) +
        xlab(paste0(names(targeted.imp[j]), " - Raw")) +
        theme(
            axis.text = element_text(size = 5), axis.title = element_text(size = 5),
            legend.text = element_text(size = 5), legend.title = element_text(size = 5)
        )

    p.log <- ggplot(targeted.imp, aes(x = log10(targeted.imp[, j]))) +
        geom_histogram(aes_string(x = log10(targeted.imp[, j])), bins = 30) +
        xlab(paste0(names(targeted.imp[j]), " - log10")) +
        theme(
            axis.text = element_text(size = 5),
            axis.title = element_text(size = 5),
            legend.text = element_text(size = 5),
            legend.title = element_text(size = 5)
        )

    p[[j]] <- ggarrange(p.raw, p.log, ncol = 2)
}

# Write pdf with all figures
# ggsave("CSF_Histograms_raw_vs_log10_KNN_targetedmetabolomics.pdf",
#        marrangeGrob(grobs=p, nrow=2, ncol=1),
#        path = dir_res)
# dev.off()
```


```{r Log Transformation, echo=FALSE, message=FALSE, warning=FALSE}
# Log10 Transform
targeted.nolog <- targeted.imp
targeted.log <- targeted.imp
targeted.log[, 23:ncol(targeted.log)] <- as.data.frame(sapply(targeted.imp[, 23:ncol(targeted.imp)], log10))
```


## Raw data distribution
```{r raw bplots, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}
bplot.raw <- cbind(sample = targeted.nolog$sample, targeted.nolog[, 23:ncol(targeted.nolog)])
bplot.raw <- bplot.raw %>%
    reshape2::melt()

print(ggplot(bplot.raw, (aes(sample, value))) +
    geom_boxplot(outlier.fill = NA) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = .5)))
```

## Log-transformed data distribution
```{r logtransf bplots, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}
bplot.log <- cbind(sample = targeted.log$sample, targeted.log[, 23:ncol(targeted.log)])
bplot.log <- bplot.log %>%
    reshape2::melt()

print(ggplot(bplot.log, (aes(sample, value))) +
    geom_boxplot(outlier.fill = NA) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = .5)))
```

## Log-transformed data distribution - without technical samples
```{r exclude technical, echo=FALSE, message=FALSE, warning=FALSE}
# edit metadata to change NAs for blank, etc
targeted.log$fullClass <- ifelse(is.na(targeted.log$Class),
    ifelse(grepl("Blank", targeted.log$sample), "technical_blank",
        ifelse(grepl("Cal", targeted.log$sample), "technical_Cal",
            ifelse(grepl("CP", targeted.log$sample), "technical_CP",
                ifelse(grepl("RPO", targeted.log$sample), "technical_RPO", NA)
            )
        )
    ), targeted.log$Class
)
targeted.log <- targeted.log %>%
    dplyr::relocate(fullClass, .after = "Cerebrospinal.fluids_Total.Tau.pg.ml")

## remove technical samples
targeted.log.f <- targeted.log %>%
    dplyr::filter(!grepl("technical", fullClass)) %>%
    dplyr::filter(!is.na(fullClass))

bplot.log <- cbind(sample = targeted.log.f$sample, targeted.log.f[, 23:ncol(targeted.log.f)])
bplot.log <- bplot.log %>%
    reshape2::melt()

print(ggplot(bplot.log, (aes(sample, value))) +
    geom_boxplot(outlier.fill = NA) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = .5)))

## remove missing metabolites (all Inf in no technical samples
inf_cols <- sapply(targeted.log.f, function(x) {
    all(is.infinite(x))
})
targeted.log.f <- targeted.log.f[, !inf_cols]
targeted.log <- targeted.log[, !inf_cols]

#### Replace Inf with 0
# Replace Inf/-Inf with 0 in columns 18 to ncol
targeted.log[, 24:ncol(targeted.log)] <- lapply(targeted.log[, 24:ncol(targeted.log)], function(col) {
    col[is.infinite(col)] <- 0
    return(col)
})

```



# PCA
```{r 1st PCA, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=6}
autoplot(prcomp(targeted.log[, 24:ncol(targeted.log)], scale. = TRUE, center = TRUE), data = targeted.log, color = "fullClass")
```

## Without technical samples
```{r 2nd PCA, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=6}
autoplot(prcomp(targeted.log.f[, 24:ncol(targeted.log.f)], scale. = TRUE, center = TRUE), data = targeted.log.f, color = "fullClass")
```

# Outliers assessment - samples
After exclusion of the technical samples, we cannot identify any outlier sample, neither in the PCA plots nor by distance to cluster center analysis.

```{r outliers on samples, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=6}
# check for outliers using a combination of PCA and kmeans clustering
autoplot(pca_all <- prcomp(targeted.log.f[, 24:ncol(targeted.log.f)], center = TRUE, scale. = TRUE), data = targeted.log.f, colour = "fullClass", x = 1, y = 2, label = TRUE, label.size = 3)
autoplot(pca_all <- prcomp(targeted.log.f[, 24:ncol(targeted.log.f)], center = TRUE, scale. = TRUE), data = targeted.log.f, colour = "fullClass", x = 1, y = 3, label = TRUE, label.size = 2)


# investigate further with k-means clustering
kmeans.result <- kmeans(targeted.log.f[, 24:ncol(targeted.log.f)], centers = 3)

# "centers" is a data frame of 3 centers but the length of iris dataset so we can canlculate distance difference easily.
centers <- kmeans.result$centers[kmeans.result$cluster, ]

# calculate distance
distances <- sqrt(rowSums((targeted.log.f[, 24:ncol(targeted.log.f)] - centers)^2))
outliers_ids <- cbind(distances, targeted.log.f)
outliers_ids2 <- cbind(kmeans.result$cluster, outliers_ids)

outliers <- outliers_ids2[order(outliers_ids2$distances), ]

# show ID of observations with the largest distance to the clusters
tibble(outliers) %>%
    dplyr::select(distances, sample) %>%
    arrange(-distances)

# all outliers with distances over 6
outliers[outliers$distances > 6, "sample"]
# targeted.log.f %>%
#    dplyr::filter(sample == outliers[outliers$distances > 6, "sample"]) %>%
#    dplyr::select(Name:fullClass) %>%
#    DT::datatable(.)

## remove outliers
# targeted.log.f <- targeted.log.f %>%
#    dplyr::filter(sample != outliers[outliers$distances > 6, "sample"]) %>%
#    as.data.frame()
```


# Outliers identification - metabolites
No differences are seen either pre- and post-winsorizing the metabolites data
```{r outliers on metabolites, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=6}
# Define outliers as anything outside the 0.01%-99.99% quantiles,
# equivalent of  4xSD away from the mean.
# Alternatively use 0.3%-99.7% equivalent of 3xSD from the mean.
tmp_outlier_range <- c(0.0001, 0.9999)

mets <- targeted.log.f[, 24:ncol(targeted.log.f)]
row.names(mets) <- targeted.log.f$sample


# Windsorizing outliers
mets_wind <- tibble(mets) %>%
    mutate(across(
        where(is.numeric),
        ~ Winsorize(., quantile(., probs = tmp_outlier_range))
    )) %>%
    data.frame()

mets_wind <- cbind(targeted.log.f[, 1:23], mets_wind)

# Visualize difference with PCA
# Principal component 1 and 2
autoplot(
    pca_all <- prcomp(targeted.log.f[, 24:ncol(targeted.log.f)],
        center = TRUE, scale. = TRUE
    ),
    data = targeted.log.f, colour = "fullClass",
    x = 1, y = 2, label = TRUE, label.size = 3
) +
    ggtitle("log-transformed metabolites PC 1 and 2")

autoplot(
    pca_all <- prcomp(mets_wind[, 24:ncol(mets_wind)],
        center = TRUE, scale. = TRUE
    ),
    data = mets_wind, colour = "fullClass",
    x = 1, y = 2, label = TRUE, label.size = 3
) +
    ggtitle("windosrized log-transformed metabolites PC 1 and 2")


# Principal component 3 and 4
autoplot(
    pca_all <- prcomp(targeted.log.f[, 24:ncol(targeted.log.f)],
        center = TRUE, scale. = TRUE
    ),
    data = targeted.log.f, colour = "fullClass",
    x = 3, y = 4, label = TRUE, label.size = 3
) +
    ggtitle("log-transformed metabolites PC 3 and 4")

autoplot(
    pca_all <- prcomp(mets_wind[, 24:ncol(mets_wind)],
        center = TRUE, scale. = TRUE
    ),
    data = mets_wind, colour = "fullClass",
    x = 3, y = 4, label = TRUE, label.size = 3
) +
    ggtitle("windosrized log-transformed metabolites PC 3 and 4")

# Truncation of outlier lipid levels did not result in any particular difference for the log-transformed data

rm(tmp_outlier_range, pca_all)
```


# Final PCA without outliers and normalized data
```{r pca2, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 10, fig.height=5}
color_key <- list(fullClass = c(
    "No cognitive impairment" = "skyblue2",
    "MCI_Non-AD" = "darkolivegreen3",
    "MCI" = "goldenrod1",
    "AD" = "salmon",
    "VaD" = "darkred"
))

mets.f <- targeted.log.f[, 24:ncol(targeted.log.f)]
row.names(mets.f) <- targeted.log.f$sample
mets.mtdt.f <- targeted.log.f[, 1:23]
row.names(mets.mtdt.f) <- targeted.log.f$sample
pca.targeted2 <- pca(as.data.frame(t(mets.f)), metadata = mets.mtdt.f)
screeplot(pca.targeted2)
```


```{r pcatools biplot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=9}
bplot <- biplot(pca.targeted2,
    showLoadings = T,
    labSize = 5, pointSize = 5, sizeLoadingsNames = 5,
    colby = "fullClass", colkey = color_key$fullClass, legendPosition = "bottom",
    encircle = FALSE, ellipse = TRUE, encircleFill = FALSE, ellipseFill = FALSE,
    lab = NULL
)
bplot
```

```{r pcatools pairs, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.width=12}
plot_legend <- get_legend(bplot)
pplot <- pairsplot(pca.targeted2, colby = "fullClass", colkey = color_key$fullClass)
print(cowplot::plot_grid(pplot, plot_legend, nrow = 2, rel_heights = c(1.9, 0.1)))
```

```{r save data, echo=FALSE, message=FALSE, warning=FALSE, eval = TRUE}
targeted.log.f$fullClass <- gsub("No cognitive impairment", "HC", targeted.log.f$fullClass)
write.table(targeted.log.f, file.path(dir_data, "CSF_targeted_logtransformed.txt"), sep = "\t", row.names = F)
```

# Cohort - confirm we have the same numbers
```{r metadata diff, echo=FALSE, message=FALSE, warning=FALSE}
metadata <- targeted.log.f[, 1:23]
metadata$Statins <- ifelse(metadata$Statins == 0, "No", "Yes")
metadata$`Hormone replacement therapy` <- ifelse(metadata$`Hormone replacement therapy` == 0, "No", "Yes")


library(tableone)
## Vector of variables to summarize
vars_test <- c(
    "Gender", "AgeAtVisit", "CognitiveSyndrome", "Diagnosis", "MMSE_Score", "Cerebrospinal.fluids_Amyloid.beta.42.pg.ml",
    "Cerebrospinal.fluids_Phospho.Tau.pg.ml", "Cerebrospinal.fluids_Total.Tau.pg.ml", "BMI", "Statins",
    "Hormone replacement therapy", "Neuropsychology_ACE_Total score", "Genetics_Risk factors_APOE", "CSF_glucose"
)
## Vector of categorical variables that need transformation
catVars <- c("Gender", "CognitiveSyndrome", "Diagnosis")

tab3 <- CreateTableOne(vars = vars_test, strata = "fullClass", data = metadata, factorVars = catVars)

tab_df <- print(tab3, printToggle = FALSE, noSpaces = TRUE)
tab_df <- tab_df[, -ncol(tab_df)]
DT::datatable(tab_df, options = list(
    scrollX = TRUE,
    pageLength = nrow(tab_df), # or -1 to display all rows if you're using lengthMenu
    lengthMenu = list(c(-1, 10, 25, 50), c("All", "10", "25", "50"))
), class = "cell-border stripe")
```

# Session
```{r session info, echo=FALSE, message=FALSE, warning=FALSE}
pander::pander(sessionInfo())
```


