---
title: "Lundbeck project - Unmetabs Plasma CGMS metabolomics - Rigshospitalet - QC"
author: "Marc Clos-Garcia"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: united
date: "`r Sys.Date()`"
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
## libraries
library(here)
library(tidyverse)
library(ggpubr)
library(factoextra)
library(VIM)
library(PCAtools)
library(ropls)
library(DT)
library(gridExtra)
library(MASS)
library(ggfortify)
library(DescTools)
library(crmn)

## directories
dir_data <- here("data/Plasma")
if (!dir.exists(dir_data)) dir.create(dir_data)
dir_res <- here("results/Plasma")
if (!dir.exists(dir_res)) dir.create(dir_res)
dir_code <- here("code")
if (!dir.exists(dir_code)) dir.create(dir_code)

color_key <- list(fullClass = c(
    "HC" = "skyblue2",
    "MCI_Non-AD" = "darkolivegreen3",
    "MCI" = "goldenrod1",
    "AD" = "salmon",
    "VaD" = "darkred",
    "technical_blank" = "pink",
    "technical_CP" = "magenta",
    "technical_RPO" = "darkmagenta"
))
```

```{r load and format datasets, echo=FALSE, message=FALSE, warning=FALSE}
#### GC-MS
metabs <- read.table(file.path(dir_data, "Area_0_20231051348.txt"), header = T, sep = "\t", skip = 4)

#### create peak column
metabs$peak <- paste0("MSpeak_", seq_len(nrow(metabs))) ## create a vector for row.names to keep an annotation object apart
metabs <- metabs %>% dplyr::select(peak, everything()) ## put the naming vector first

#### extract annotation and abundance matrices
met_annotation <- metabs[, 1:27]
metabs <- metabs[, c(1, 30:ncol(metabs))]

## transpose and convert to DF
row.names(metabs) <- metabs$peak
metabs$peak <- NULL
metabs <- as.data.frame(t(metabs))

###### extract sample numbers
row.names(metabs) <- gsub("_backup", "", row.names(metabs))

metabs <- metabs %>%
    tibble::rownames_to_column("Data File") %>%
    tidyr::separate("Data File", sep = "_", into = c("P_number", "P_name", "Date", "Run", "sample", "vial", "injection"), remove = FALSE) %>%
    dplyr::mutate(sample = ifelse(is.na(sample), `Data File`, sample)) %>%
    dplyr::select(-P_number, -P_name, -Date, -vial, -injection) %>%
    as.data.frame()

# write.table(met_annotation, file.path(dir_data, "mets_annotationfile.txt"), sep = "\t")

#### clinical - sample IDs
mtdt <- read.delim(file.path(dir_data, "Sample codenumbers plus data 2023_01_03_KH.csv"), header = T, sep = ",")

#### IDs matching file
matching <- read.table(file.path(dir_data, "0111_Lundbeck_QQQ_plasma_STENO_ID_DDBBno_each_sample.csv"), header = T, sep = ",", skip = 1)
matching <- matching[, -3]
```


```{r add DDBBno ids to metadata and merge all files, echo=FALSE, message=FALSE, warning=FALSE}
#### add ID to metadata
mtdt <- mtdt %>%
    dplyr::left_join(matching, by = "DDBBno") %>%
    dplyr::select(STENO.ID, DDBBno, everything())
mtdt$STENO.ID <- as.character(mtdt$STENO.ID)

#### add extra metadata
extra_metadata <- readxl::read_excel(file.path(dir_data, "RH_Clinical info BMI_Statin_HRT_2025_06_23.xlsx"), sheet = 1, col_names = T)
extra_metadata <- extra_metadata %>% dplyr::select(DDBBno, BMI, Statins, "Hormone replacement therapy", `Neuropsychology_ACE_Total score`, `Genetics_Risk factors_APOE`, CSF_glucose = "Cerebrospinal fluids_Glucose#mmol/l")

mtdt <- mtdt %>% 
  dplyr::left_join(extra_metadata, by = "DDBBno") %>% 
  dplyr::mutate(Statins = ifelse(Statins == "Nej", 0, 1)) %>% 
  dplyr::mutate("Hormone replacement therapy" = ifelse(`Hormone replacement therapy` == "Nej", 0, 1))

#### merge metabolomics with metadata
metabs <- metabs %>%
    dplyr::left_join(mtdt, by = c("sample" = "STENO.ID")) %>%
    dplyr::select(
        Name = `Data File`, sample, DDBBno, Gender, AgeAtVisit, CognitiveSyndrome,
        Diagnosis, Class, MMSE_Score, Cerebrospinal.fluids_Amyloid.beta.42.pg.ml,
        Cerebrospinal.fluids_Phospho.Tau.pg.ml, Cerebrospinal.fluids_Total.Tau.pg.ml, BMI, 
        Statins, "Hormone replacement therapy", "Neuropsychology_ACE_Total score", 
        "Genetics_Risk factors_APOE", CSF_glucose, everything())

## clean up
rm(matching)
```


```{r data transformations, echo=FALSE, message=FALSE, warning=FALSE}
## data transformations
### transform to numbers
metabs[, 20:ncol(metabs)] <- sapply(metabs[, 20:ncol(metabs)], as.numeric) ## convert to numeric
```

# Context
In this report, we perform a quality-control evaluation of the metabs metabolomics dataset. In total, the dataset included up to `r ncol(metabs[, 17:ncol(metabs)])` annotated metabs metabolites. The process presented here includes:

* Evaluating the ratios between blanks and samples
* Evaluation of Coefficient of Variation (CV) in the sample pools
* Identification and processing of duplicate samples
* Missing values evaluation and imputation
* Data normalization
* Identification and removal of outliers


## How many annotated and unannotated peaks?
```{r evaluate annotated, echo=FALSE, warning=FALSE, message=FALSE}
annot_summary <- as.data.frame(table(ifelse(met_annotation$Metabolite.name == "Unknown", "Unannotated", "Annotated")))
annot_summary %>%
    dplyr::select(Classification = Var1, n = Freq) %>%
    dplyr::mutate(percentage = (n / sum(n) * 100)) %>%
    dplyr::mutate_if(is.numeric, round, 2) %>%
    DT::datatable(.)
```

In total, we see `r annot_summary[1, 2]` annotated metabolites and `r annot_summary[2, 2]`

# Results {.tabset}

## Annotated peaks
```{r keep annotated mets, echo=FALSE, message=FALSE, warning=FALSE}
# filter out unannotated peaks from annotation file
annotated_mets <- met_annotation[met_annotation$Metabolite.name != "Unknown", ]

# keep only annotated metabolites
metabs.annot <- metabs[, names(metabs) %in% annotated_mets$peak]
metabs <- cbind(metabs[, 1:19], metabs.annot)
```


### Check blanks vs other samples metabolites
Looking at the metabolite average abundance per sample type, we see that only 4 metabolites not being present in blanks: *linoleic acid, oleic acid, stearic acid* and *threonine*.

Ideally, we would like to keep only those metabolites not present in the blank samples or, at least, present in 3x times in the samples than in the blanks. 

```{r average abundance comparison -annotated-, echo=FALSE, message=FALSE, warning=FALSE}
blanks <- metabs %>% dplyr::filter(grepl("blank", sample))
RPO <- metabs %>% dplyr::filter(grepl("RPO", sample))
CP <- metabs %>% dplyr::filter(grepl("CP", sample))

samples <- metabs %>% dplyr::filter(!(sample %in% c(blanks$sample, RPO$sample, CP$sample)))

## compute mean values per metabolite
blanks.avg <- sapply(blanks[, 20:ncol(blanks)], median, na.rm = T)
samples.avg <- sapply(samples[, 20:ncol(samples)], median, na.rm = T)
RPO.avg <- sapply(RPO[, 20:ncol(RPO)], median, na.rm = T)
CP.avg <- sapply(CP[, 20:ncol(CP)], median, na.rm = T)

avg.abundance <- data.frame(
    "blanks" = blanks.avg,
    "RPO" = RPO.avg,
    "CP" = CP.avg,
    "samples" = samples.avg
)

avg.abundance %>%
    tibble::rownames_to_column("metabolite") %>%
    reshape2::melt() %>%
    ggplot(aes(metabolite, log10(value), group = variable, fill = variable)) +
    geom_col(position = position_dodge()) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))

avg.abundance %>%
    tibble::rownames_to_column("metabolite") %>%
    reshape2::melt() %>%
    dplyr::filter(metabolite %in% sample(unique(metabolite), 15)) %>% 
    ggplot(aes(metabolite, log10(value), group = variable, fill = variable)) +
    geom_col(position = position_dodge()) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
```

```{r FC blanks-samples, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10}

FC_peaks <- as.data.frame(sapply(samples[, 20:ncol(samples)], median, na.rm=T)/sapply(blanks[, 20:ncol(samples)], median, na.rm=T)) %>% 
  tibble::rownames_to_column("peak") 
names(FC_peaks) <- c("peak", "FC")  
FC_peaks$color <- ifelse(log2(FC_peaks$FC)>0, "black", "red")
ggplot(FC_peaks, aes(log2(FC), reorder(peak, FC), fill = color)) + geom_col() + 
  geom_vline(xintercept = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  scale_fill_manual(values=c("black", "darkred")) +
  guides(fill=F)

# FC_peaks %>% 
#   mutate(log2FC = log2(FC)) %>% 
#   dplyr::filter(log2FC > 1)

metabs.remove <- FC_peaks %>% 
  mutate(log2FC = log2(FC)) %>% 
  dplyr::filter(log2FC < 0)


standards <- met_annotation[grepl("Steno", met_annotation$Metabolite.name), ]$peak
metabs.remove <- metabs.remove %>% dplyr::filter(!peak %in% standards)

metabs <- metabs[, !names(metabs) %in% metabs.remove$peak]
```


```{r subtract blanks -carryover- from areas}
mspeak_cols <- grep("^MSpeak_", names(metabs), value = TRUE)
blanks.avg <- blanks.avg[names(blanks.avg) %in% mspeak_cols]

metabs.subtracted <- metabs
metabs.subtracted[, mspeak_cols] <- 
  sweep(metabs[, mspeak_cols], 2, blanks.avg, FUN = "-")

metabs.subtracted[metabs.subtracted < 0] <- NA

metabs <- metabs.subtracted
```

### Look at pooled samples - RPO
If we evaluate the Coefficient of Variation (CV) for the Real Pool (RPO) samples, we only identify *cysteine* as having a large CV (>30%), which would result in its removal from the analyses.

```{r CV RPO -annotated-, echo=FALSE, message=FALSE, warning=FALSE}
# RPO %>%
#     dplyr::select(Run, row.names(avg.abundance)) %>%
#     reshape2::melt() %>%
#     ggplot(aes(variable, log10(value))) +
#     geom_boxplot()


cv <- function(x) {
    cv <- sd(x) / abs(mean(x)) * 100
    return(cv)
}

cv.rpo <- sapply(RPO[, row.names(avg.abundance)], cv)
cv.rpo <- data.frame(Metabolite = names(cv.rpo), CV = cv.rpo)
cv.rpo <- cv.rpo %>%
    arrange(CV) %>%
    dplyr::mutate(Metabolite = factor(Metabolite, levels = Metabolite))
ggplot(cv.rpo, aes(CV, Metabolite)) +
    geom_col() +
    geom_vline(xintercept = 30, color = "red") +
    labs(
        title = "RPO pools Coefficient of Variation",
        x = "Coefficient of Variation (%)"
    )
```

### Look at pooled samples - CP
If we evaluate the Coefficient of Variation (CV) for the Clinical Pool (CP) samples, we see up to 4 metabolites: *cysteine, stearic acid, arachidonic acid* and *alpha ketoglutarate* as having a large CV (>30%), which would result in their removal from the analyses.

```{r CV CP -annotated-, echo=FALSE, message=FALSE, warning=FALSE}
# CP %>%
#     dplyr::select(Run, row.names(avg.abundance)) %>%
#     reshape2::melt() %>%
#     ggplot(aes(variable, log10(value))) +
#     geom_boxplot()


cv.cp <- sapply(CP[, row.names(avg.abundance)], cv)
cv.cp <- data.frame(Metabolite = names(cv.cp), CV = cv.cp)
cv.cp <- cv.cp %>%
    arrange(CV) %>%
    dplyr::mutate(Metabolite = factor(Metabolite, levels = Metabolite))
ggplot(cv.cp, aes(CV, Metabolite)) +
    geom_col() +
    geom_vline(xintercept = 30, color = "red") +
    labs(
        title = "CP pools Coefficient of Variation",
        x = "Coefficient of Variation (%)"
    )
```


### Check for duplicates {.tabset}
```{r duplicates -annotated-, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# sort(table(metabs$sample), decreasing = TRUE) ## we check duplicates by sample name, no duplicates are observed except for blank17, CP17 and RPO17

# Check missing patterns to see if any of them has more missing data
# This is also a type of QCing the data to check how well duplicates correlate
dups.df <- metabs[duplicated(metabs$sample), ]
datatable(dups.df)

dup.samples <- metabs[duplicated(metabs$sample), "sample"]

for (x in dup.samples) {
    cat("#### ", x, "\n\n")
    nas <- data.frame(cbind(
        summary(rowSums(is.na(metabs[which(metabs$sample == x)[1], 20:ncol(metabs)]))),
        summary(rowSums(is.na(metabs[which(metabs$sample == x)[2], 20:ncol(metabs)])))
    ))
    names(nas) <- which(metabs$sample == x)
    print(htmltools::tagList(DT::datatable(nas,
        escape = T, height = "100%", class = "cell-border stripe",
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
    cat("\n\n")
    # Plot the two pairs of data
    plot(t(metabs[which(metabs$sample == x), 20:ncol(metabs)]))
    cat("\n\n")
}

## samples look very similar, we'll remove one duplicate randomly
metabs <- metabs[!duplicated(metabs$sample), ]

# clean up
rm(dup.samples, nas)
```


### {.unnumbered .unlisted}
We see `r nrow(dups.df)` duplicates. After evaluating the metabolites distribution across these duplicates, we decide to randomly remove one duplicate for each pair.

### Missing values
```{r NAs -annotated-, echo=FALSE, message=FALSE, warning=FALSE}
# Check participants with all met data missing
metabs$allNA <- apply(metabs[, 20:ncol(metabs)],
    MARGIN = 1, FUN = function(x) all(is.na(x))
)

summary(metabs$allNA)

# NONE missing- remove allNA column
metabs$allNA <- NULL


# check missingess per individual
summary(rowSums(is.na(metabs[, 20:ncol(metabs)])))
hist(summary(rowSums(is.na(metabs[, 20:ncol(metabs)]))), breaks = 10)
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
# 0.0000  0.0000  0.0000  0.3636  0.0000  6.0000

row.names(metabs) <- metabs$sample


## Exclude individuals with missing data in more than 30% of metabolites -!!!
metabs.f <- metabs[, 20:ncol(metabs)][which(rowMeans(is.na(metabs[, 20:ncol(metabs)])) <= 0.3), ]

# 0 individuals have more than 20% of missing data-> 297 individuals left
# Exclude metabolites with missing data in more than 30% of metabolites
metabs.f2 <- metabs.f[, which(colMeans(is.na(metabs.f)) <= 0.3), ]

dim(metabs.f2)
# [1]  297 30  left
# no metabolites missing more than 20%

## merge back the filtered dataset
#### keep standards
stds.peaks <- met_annotation[grepl("Steno", met_annotation$Metabolite.name), ]$peak
stds <- metabs.f[, stds.peaks]

metabs.f <- merge(metabs[, 1:19], metabs.f2, by = "row.names")
metabs.f$Row.names <- NULL
dim(metabs.f)
```

No samples were removed due to the number of missing values (all <20%). 
Missing values were imputed using kNN clustering approach with k=10.

```{r impute missing values -annotated-, echo=FALSE, message=FALSE, warning=FALSE}
### add back standards (for normalization/plotting)
stds.toadd <- setdiff(standards, colnames(metabs.f))
metabs.stds <- metabs[, stds.toadd]
metabs.stds <- metabs.stds[row.names(metabs.stds) %in% metabs.f$sample, ]
metabs.stds <- metabs.stds[match(metabs.f$sample, row.names(metabs.stds)), ]
metabs.f <- cbind(metabs.f, metabs.stds)
  
### impute missing values with kNN
metabs.imp <- kNN(metabs.f[, 20:ncol(metabs.f)], k = 10)
metabs.imp <- cbind(metabs.f[, 1:19], metabs.imp)
metabs.imp <- metabs.imp %>%
    dplyr::select(-contains("_imp")) %>%
    as.data.frame()
metabs.imp[, 20:ncol(metabs.imp)] <- sapply(metabs.imp[, 20:ncol(metabs.imp)], as.numeric) # convert metabolite data back to numeric

rm(metabs, metabs.f)
```


### Data normalization - internal standards
```{r standards normalization, echo=FALSE, message=FALSE, warning=FALSE}
# edit metadata to change NAs for blank, etc
metabs.imp$fullClass <- ifelse(is.na(metabs.imp$Class),
    ifelse(grepl("lank", metabs.imp$sample), "technical_blank",
        ifelse(grepl("Cal", metabs.imp$sample), "technical_Cal",
            ifelse(grepl("CP", metabs.imp$sample), "technical_CP",
                ifelse(grepl("RPO", metabs.imp$sample), "technical_RPO", NA)
            )
        )
    ), metabs.imp$Class
)
metabs.imp <- metabs.imp %>%
    dplyr::relocate(fullClass, .after = "Cerebrospinal.fluids_Total.Tau.pg.ml")
metabs.imp$fullClass <- gsub("No cognitive impairment", "HC", metabs.imp$fullClass)

## remove empty samples
metabs.imp <- metabs.imp[!is.na(metabs.imp$fullClass), ]

metabolites <- metabs.imp
row.names(metabolites) <- metabolites$sample

metadata <- metabolites[, 1:20]
metabolites <- metabolites[, 21:ncol(metabolites)]
## filter out annotated_mets
annotated_mets <- annotated_mets %>% dplyr::filter(peak %in% names(metabolites))
# identical(names(metabolites), annotated_mets$peak)

colnames(metabolites) <- make.names(annotated_mets$Metabolite.name)

## list standards
# Below internal standards were added to each sample before the extraction. So they could be used to normalize individual compound.
before_process <- c(
    "Glutamic acid d5 3TMS ISTD Steno", "Valine d8 2TMS ISTD Steno", "Heptadecanoic acid d33 TMS ISTD Steno",
    "Succinic acid d4 2TMS ISTD Steno", "Tritriacontane C33 ISTD Steno"
)
before_process <- make.names(before_process)

# Below internal standards were added to each sample during derivatization. These internal standards could be used to monitor the % RSD of these compound across the batch analysis.
derivatization <- c(
    "Undecane C11 ISTD Steno", "Pentadecane C15 ISTD Steno", "Heptadecane C17 ISTD Steno",
    "Heneicosane C21 ISTD Steno", "Pentacosane C25 Steno ISTD", "Nonacosane C29 ISTD Steno"
)
derivatization <- make.names(derivatization)

# "X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno"

## normalize
metabolites.t <- t(metabolites)
is_std <- rownames(metabolites.t) %in% before_process
G <- with(metadata, model.matrix(~ -1 + fullClass))

crmn.norm <- normalize(metabolites.t, "crmn", factors = G, standards = is_std)
nomis.norm <- normalize(metabolites.t, "nomis", standards = is_std)

pca.crmn <- pca(scale(log(t(crmn.norm))))
pca.crmn <- scores(pca.crmn)
pca.crmn <- pca.crmn |>
    as.data.frame() |>
    tibble::rownames_to_column("sample") |>
    left_join(metadata, by = "sample")

pca.nomis <- pca(scale(log(t(nomis.norm))))
pca.nomis <- scores(pca.nomis)
pca.nomis <- pca.nomis |>
    as.data.frame() |>
    tibble::rownames_to_column("sample") |>
    left_join(metadata, by = "sample")

ggpubr::ggarrange(
    ggplot(pca.crmn, aes(PC1, PC2, color = Run, shape = fullClass)) +
        geom_point() +
        labs(title = "CRMN") +
        guides(color = F) +
        scale_shape_manual(values = c(15, 16, 17, 18, 25, 8, 4, 9)),
    ggplot(pca.nomis, aes(PC1, PC2, color = Run, shape = fullClass)) +
        geom_point() +
        labs(title = "NOMIS") +
        guides(color = F) +
        scale_shape_manual(values = c(15, 16, 17, 18, 25, 8, 4, 9)),
    common.legend = TRUE
)



# mets.norm <- as.data.frame(t(mets.norm))
# row.names(mets.norm) == metadata$sample
```

#### 1. Variance Partitioning (R² for technical vs biological factors)

NOMIS: Higher variance explained by technical factors → means technical effects still dominate after normalization.

CRMN: Much lower variance explained by technicals → suggests NOMIS reduces technical bias more effectively. ✅

```{r var partition normalization approaches, echo=FALSE, message=FALSE, warning=FALSE}
metadata$type <- ifelse(metadata$fullClass %in% c("HC", "MCI_Non-AD", "MCI", "AD", "VaD"), "sample",
    ifelse(grepl("blank", metadata$fullClass), "blank", "QC")
)
# Example function to compute R² from a linear model
get_r2 <- function(y, metadata) {
    model <- lm(y ~ fullClass + type, data = metadata)
    summary(model)$adj.r.squared
}

# Apply across features
r2_crmn <- apply(crmn.norm, 1, function(x) get_r2(x, metadata))
r2_nomis <- apply(nomis.norm, 1, function(x) get_r2(x, metadata))

# Compare distributions
boxplot(r2_crmn, r2_nomis,
    names = c("CRMN", "NOMIS"),
    main = "Variance explained (adj. R²)"
)
```

#### 2. Coefficient of Variation (CV) for QC Samples

NOMIS: Slightly higher spread in CVs.

CRMN: Lower median and tighter CVs in QC samples → better reproducibility. ✅

```{r cv normalizations, echo=FALSE, message=FALSE, warning=FALSE}
# Subset QC samples
qc_samples <- metadata %>% filter(type == "QC")

# Calculate CV
cv_fun <- function(mat, qc_ids) {
    mat_qc <- mat[, qc_ids]
    apply(mat_qc, 1, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) * 100)
}

cv_crmn <- cv_fun(crmn.norm, qc_samples$sample)
cv_nomis <- cv_fun(nomis.norm, qc_samples$sample)

# Plot comparison
boxplot(cv_crmn, cv_nomis,
    names = c("CRMN", "NOMIS"),
    main = "QC Coefficient of Variation (%)"
)
```

#### 3. Relative Log Expression (RLE) Plots

Similar distributions, lower distributions for blanks

```{r RLE plots, echo=FALSE, message=FALSE, warning=FALSE}
library(EDASeq)

par(mfrow = c(1, 2))
plotRLE(crmn.norm, outline = FALSE, main = "CRMN RLE", ylab = "Relative Log Expression")
plotRLE(nomis.norm, outline = FALSE, main = "NOMIS RLE", ylab = "Relative Log Expression")
```


#### 4. Clustering of Replicates

CRMN clusters pooled samples and blanks separately from the samples.

```{r clustering heatmap, echo=FALSE, message=FALSE, warning=FALSE}
library(pheatmap)

# Distance matrix and clustering
dist_crmn <- dist(t(crmn.norm))
dist_nomis <- dist(t(nomis.norm))

pheatmap(as.matrix(dist_crmn), main = "CRMN - Sample Clustering")
pheatmap(as.matrix(dist_nomis), main = "NOMIS - Sample Clustering")
```

#### 5. Biological Signal Preservation (Example: AD vs HC)
Both methods recover similar biological hits.

```{r LMs, echo=FALSE, message=FALSE, warning=FALSE}
# Example differential test
library(limma)

# Design matrix
metadata$fullClass2 <- factor(make.names(metadata$fullClass), levels = make.names(unique(metadata$fullClass)))
design <- model.matrix(~ 0 + metadata$fullClass2)
colnames(design) <- levels(metadata$fullClass2)

contrast.matrix <- makeContrasts(
    ADvsHC = AD - HC,
    levels = design
)

# Fit linear model
fit_crmn <- lmFit(crmn.norm, design)
fit_crmn <- contrasts.fit(fit_crmn, contrast.matrix)
fit_crmn <- eBayes(fit_crmn)

fit_nomis <- lmFit(nomis.norm, design)
fit_nomis <- contrasts.fit(fit_nomis, contrast.matrix)
fit_nomis <- eBayes(fit_nomis)

# Get top results
top_crmn <- topTable(fit_crmn, coef = "ADvsHC", number = 20)
top_nomis <- topTable(fit_nomis, coef = "ADvsHC", number = 20)

head(top_crmn)
head(top_nomis)
```




```{r center and scale data, echo=FALSE, message=FALSE, warning=FALSE}
# mets.norm <- nomis.norm
mets.norm <- crmn.norm

mets.norm <- log10(mets.norm)
mets.norm <- as.data.frame(t(mets.norm))
mets.scaled <- as.data.frame(sapply(mets.norm, function(x) {
    scale(x, center = TRUE, scale = TRUE)
}))
row.names(mets.scaled) <- row.names(mets.norm)
normalized_data <- cbind(metadata, mets.scaled)
# write.table(normalized_data, file.path(dir_data, "untargetedANNOT_NOMISnomralized.txt"), row.names = FALSE, sep = "\t")
write.table(normalized_data, file.path(dir_data, "untargetedANNOT_CRMNnomralized.txt"), row.names = FALSE, sep = "\t")
```

```{r Visualize distribution 1 -annotated-, echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE}
colnames(metabs.imp)[23:ncol(metabs.imp)] <- make.names(annotated_mets$Metabolite.name)


# Vector of all lipids to iterate through
Metabolite_names <- names(normalized_data)[23:ncol(normalized_data)]
# Visualize histograms
p <- list()
for (j in Metabolite_names) {
    p.raw <- ggplot(metabs.imp, aes(x = (metabs.imp[, j]))) +
        geom_histogram(aes_string(x = metabs.imp[, j]), bins = 30) +
        xlab(paste0(names(metabs.imp[j]), " - Raw")) +
        theme(
            axis.text = element_text(size = 5), axis.title = element_text(size = 5),
            legend.text = element_text(size = 5), legend.title = element_text(size = 5)
        )

    p.log <- ggplot(normalized_data, aes(x = normalized_data[, j])) +
        geom_histogram(aes_string(x = normalized_data[, j]), bins = 30) +
        xlab(paste0(names(normalized_data[j]), " - log10")) +
        theme(
            axis.text = element_text(size = 5),
            axis.title = element_text(size = 5),
            legend.text = element_text(size = 5),
            legend.title = element_text(size = 5)
        )

    p[[j]] <- ggarrange(p.raw, p.log, ncol = 2)
}

# Write pdf with all figures
# ggsave("Plasma_Histograms_raw_vs_CRMN_log10_KNN_metabsmetabolomics.pdf",
#        marrangeGrob(grobs=p, nrow=2, ncol=1),
#        path = dir_res)
# dev.off()
```


#### Normalized data distribution 
```{r norm plots -annotated-, echo=FALSE, message=FALSE, warning=FALSE}
bplot.log <- cbind(sample = normalized_data$sample, normalized_data[, 21:ncol(normalized_data)])
bplot.log <- bplot.log %>%
    reshape2::melt()

print(ggplot(bplot.log, (aes(sample, value))) +
    geom_boxplot(outlier.fill = NA) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = .5)))
```

#### PCA before removing technical samples
```{r PCA norm, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=6.5}
metabolites <- normalized_data[, 23:ncol(normalized_data)]
metadata <- normalized_data[, 1:21]

pca.normalized.all <- PCAtools::pca(t(metabolites), metadata)
bplot <- biplot(pca.normalized.all,
    showLoadings = T,
    labSize = 5, pointSize = 5, sizeLoadingsNames = 5,
    colby = "fullClass", colkey = color_key$fullClass, legendPosition = "bottom",
    encircle = FALSE, ellipse = TRUE, encircleFill = FALSE, ellipseFill = FALSE,
    lab = NULL
)
bplot
```

#### Log-transformed data distribution - without technical samples
```{r exclude technical -annotated-, echo=FALSE, message=FALSE, warning=FALSE}
# edit metadata to change NAs for blank, etc

## remove technical samples
normalized_data <- normalized_data %>%
    dplyr::filter(!grepl("technical", fullClass)) %>%
    dplyr::filter(!is.na(fullClass))

bplot.log <- cbind(sample = normalized_data$sample, normalized_data[, 23:ncol(normalized_data)])
bplot.log <- bplot.log %>%
    reshape2::melt()

print(ggplot(bplot.log, (aes(sample, value))) +
    geom_boxplot(outlier.fill = NA) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = .5)))
```


### Check derivatization and injector standards distribution

#### Derivatization standards

```{r stds distribution, echo=FALSE, message=FALSE, warning=FALSE}
met_annotation_stds <- met_annotation[met_annotation$peak %in% names(stds), ]
# identical(names(stds), met_annotation_stds$peak)
colnames(stds) <- make.names(met_annotation_stds$Metabolite.name)

derivatization <- c(derivatization)
injector <- "X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno"

stds.df <- normalized_data[, derivatization]

stds.df$sample <- normalized_data$sample
stds.df.m <- reshape2::melt(stds.df)
derv.plot1 <- ggplot(stds.df.m, aes(variable, value)) +
    geom_boxplot() +
    theme_bw() +
    labs(x = "Derivatization standards", y = "Normalized expression", title = "Derivatization standards BEFORE correction") +
    theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1))
```

```{r correct deriv, echo=FALSE, message=FALSE, warning=FALSE}
## compute median per sample (in derivatization subset)
stds.df$sample <- NULL
median_std <- apply(stds.df, 1, median, na.rm = TRUE)
median_std <- as.numeric(median_std)

# Scale so median factor = 0 (centered on log scale)
median_std <- median_std - median(median_std, na.rm = T)

# Apply normalization by subtraction (since we're in log10 space)
metabolites.names <- setdiff(colnames(normalized_data[, 23:ncol(normalized_data)]), derivatization)
metabolites_norm_derv <- normalized_data
# metabolites_norm_derv[, metabolites.names] <- sweep(mets.norm[, metabolites.names], 1, median_std, "-")
metabolites_norm_derv[, 23:ncol(metabolites_norm_derv)] <- sweep(normalized_data[, 23:ncol(normalized_data)], 1, median_std, "-")

## re-check dervs
derv.plot2 <- metabolites_norm_derv %>%
    dplyr::select(derivatization) %>%
    reshape2::melt() %>%
    ggplot(aes(variable, value)) +
    geom_boxplot() +
    theme_bw() +
    labs(x = "Derivatization standards", y = "Normalized expression", title = "Derivatization standards AFTER correction") +
    theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1))
```


```{r}
ggpubr::ggarrange(derv.plot1, derv.plot2)


## remove standards from dataset
normalized_data <- normalized_data[, !(names(normalized_data) %in% c(before_process, derivatization))]
# write.table(normalized_data, file.path(dir_data, "CSF_untargetedALL_standards_normalized_metabolites"), row.names = FALSE, sep = "\t")

metabs.log <- normalized_data
```


### Check injector standard
```{r std injection, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}
# Compute mean & SD on log10 scale
mean_injector <- mean(normalized_data$X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno, na.rm = TRUE)
sd_injector <- sd(normalized_data$X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno, na.rm = TRUE)

# Flag samples more than 2 SD from mean (adjust threshold if needed)
injector_outliers <- ifelse(abs(normalized_data$X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno - mean_injector) > 3 * sd_injector, "yes", "no")



normalized_data %>%
    dplyr::select(sample, injector, fullClass) %>%
    dplyr::mutate(outlier = injector_outliers) %>%
    dplyr::filter(!is.na(outlier)) %>% 
    ggplot(aes(sample, X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno, shape = outlier, color = fullClass)) +
    geom_point() +
    geom_hline(yintercept = mean(normalized_data[, injector]), color = "red") +
    geom_hline(yintercept = mean(normalized_data[, injector]) + 3 * sd_injector, color = "red", linetype = "dashed") +
    geom_hline(yintercept = mean(normalized_data[, injector]) - 3 * sd_injector, color = "red", linetype = "dashed") +
    scale_color_manual(values = color_key$fullClass) +
    theme_bw() +
    labs(x = NULL, title = "Injector values across samples after normalization") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
```

We see some outliers outside of the ±3*SD threshold. We also see that the class groups are evenly distributed along the injector values. 


```{r injector outliers, echo=FALSE, message=FALSE, warning=FALSE}
normalized_data %>%
    dplyr::mutate(outlier = injector_outliers) %>%
    dplyr::filter(outlier == "yes") %>%
    dplyr::select(
        Name, sample, DDBBno, Gender, AgeAtVisit, CognitiveSyndrome, Diagnosis,
        Class, fullClass, MMSE_Score, outlier, Cerebrospinal.fluids_Amyloid.beta.42.pg.ml,
        Cerebrospinal.fluids_Phospho.Tau.pg.ml, Cerebrospinal.fluids_Total.Tau.pg.ml, 
        "BMI", "Statins", "Hormone replacement therapy", "Neuropsychology_ACE_Total score", 
        "Genetics_Risk factors_APOE", "CSF_glucose"
    ) %>%
    DT::datatable(., options = list(scrollX = TRUE))


normalized_data <- normalized_data %>%
    dplyr::mutate(outlier = injector_outliers) %>%
    dplyr::filter(outlier != "yes") %>%
    dplyr::select(-X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno, -outlier) %>%
    as.data.frame()

## remove standards from dataset
normalized_data <- normalized_data[, !(names(normalized_data) %in% c(before_process, derivatization, "X4.4..Dibromooctafluorobiphenyl..Injector.reproducibility..Steno"))]

# write.table(normalized_data, file.path(dir_data, "untargetedALL_standards_normalized_metabolites"), row.names = FALSE, sep = "\t")

metabs.log <- normalized_data
```


### PCAs
```{r 1st PCA -annotated-, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=6}
pca.normalized.f <- PCAtools::pca(t(metabs.log[, 23:ncol(metabs.log)]), metadata = metabs.log[, 1:21])
bplot <- biplot(pca.normalized.f,
    showLoadings = T,
    labSize = 5, pointSize = 5, sizeLoadingsNames = 5,
    colby = "fullClass", colkey = color_key$fullClass, legendPosition = "bottom",
    encircle = FALSE, ellipse = TRUE, encircleFill = FALSE, ellipseFill = FALSE,
    lab = NULL
)
bplot
# autoplot(prcomp(metabs.log[, 17:ncol(metabs.log)]), data = metabs.log, color = "fullClass")
```

### Outliers assessment - samples
After exclusion of the technical samples, we cannot identify any outlier sample, neither in the PCA plots nor by distance to cluster center analysis.

```{r outliers on samples -annotated-, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=6}
# check for outliers using a combination of PCA and kmeans clustering
autoplot(pca_all <- prcomp(metabs.log[, 23:ncol(metabs.log)]), data = metabs.log, colour = "fullClass", x = 1, y = 2, label = TRUE, label.size = 3)
autoplot(pca_all <- prcomp(metabs.log[, 23:ncol(metabs.log)]), data = metabs.log, colour = "fullClass", x = 1, y = 3, label = TRUE, label.size = 2)


# investigate further with k-means clustering
kmeans.result <- kmeans(metabs.log[, 23:ncol(metabs.log)], centers = 2)

# "centers" is a data frame of 3 centers but the length of iris dataset so we can canlculate distance difference easily.
centers <- kmeans.result$centers[kmeans.result$cluster, ]

# calculate distance
distances <- sqrt(rowSums((metabs.log[, 23:ncol(metabs.log)] - centers)^2))
outliers_ids <- cbind(distances, metabs.log)
outliers_ids2 <- cbind(kmeans.result$cluster, outliers_ids)

outliers <- outliers_ids2[order(outliers_ids2$distances), ]

# show ID of observations with the largest distance to the clusters
tibble(outliers) %>%
    dplyr::select(distances, sample) %>%
    arrange(-distances)

# all outliers with distances over 14
outliers[outliers$distances > 14, "sample"]
metabs.log %>%
    dplyr::filter(sample %in% outliers[outliers$distances > 14, "sample"]) %>%
    dplyr::select(Name:fullClass) %>%
    DT::datatable(.)

## remove outliers
metabs.log <- metabs.log %>%
    dplyr::filter(!sample %in% outliers[outliers$distances > 14, "sample"]) %>%
    as.data.frame()
```


### Outliers identification - metabolites
No differences are seen either pre- and post-winsorizing the metabolites data
```{r outliers on metabolites -annotated-, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=6}
# Define outliers as anything outside the 0.01%-99.99% quantiles,
# equivalent of  4xSD away from the mean.
# Alternatively use 0.3%-99.7% equivalent of 3xSD from the mean.
tmp_outlier_range <- c(0.0001, 0.9999)

mets <- metabs.log[, 23:ncol(metabs.log)]
row.names(mets) <- metabs.log$sample


# Windsorizing outliers
mets_wind <- tibble(mets) %>%
    mutate(across(
        where(is.numeric),
        ~ Winsorize(., quantile(., probs = tmp_outlier_range))
    )) %>%
    data.frame()

mets_wind <- cbind(metabs.log[, 1:22], mets_wind)

# Visualize difference with PCA
# Principal component 1 and 2
autoplot(
    pca_all <- prcomp(metabs.log[, 23:ncol(metabs.log)],
        center = TRUE, scale. = TRUE
    ),
    data = metabs.log, colour = "fullClass",
    x = 1, y = 2, label = TRUE, label.size = 3
) +
    ggtitle("log-transformed metabolites PC 1 and 2")

autoplot(
    pca_all <- prcomp(mets_wind[, 23:ncol(mets_wind)],
        center = TRUE, scale. = TRUE
    ),
    data = mets_wind, colour = "fullClass",
    x = 1, y = 2, label = TRUE, label.size = 3
) +
    ggtitle("windosrized log-transformed metabolites PC 1 and 2")


# Principal component 3 and 4
autoplot(
    pca_all <- prcomp(metabs.log[, 23:ncol(metabs.log)],
        center = TRUE, scale. = TRUE
    ),
    data = metabs.log, colour = "fullClass",
    x = 3, y = 4, label = TRUE, label.size = 3
) +
    ggtitle("log-transformed metabolites PC 3 and 4")

autoplot(
    pca_all <- prcomp(mets_wind[, 23:ncol(mets_wind)],
        center = TRUE, scale. = TRUE
    ),
    data = mets_wind, colour = "fullClass",
    x = 3, y = 4, label = TRUE, label.size = 3
) +
    ggtitle("windosrized log-transformed metabolites PC 3 and 4")

# Truncation of outlier lipid levels did not result in any particular difference for the log-transformed data

rm(tmp_outlier_range, pca_all)
```


### Final PCA without outliers and normalized data
```{r pca2 -annotated-, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 10, fig.height=5}
mets.f <- metabs.log[, 23:ncol(metabs.log)]
row.names(mets.f) <- metabs.log$sample
mets.mtdt.f <- metabs.log[, 1:21]
row.names(mets.mtdt.f) <- metabs.log$sample
pca.metabs2 <- PCAtools::pca(as.data.frame(t(mets.f)), metadata = mets.mtdt.f)
screeplot(pca.metabs2, components = 1:25)
```


```{r pcatools biplot -annotated-, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=9}
bplot <- biplot(pca.metabs2,
    showLoadings = T,
    labSize = 5, pointSize = 5, sizeLoadingsNames = 5,
    colby = "fullClass", colkey = color_key$fullClass, legendPosition = "bottom",
    encircle = FALSE, ellipse = TRUE, encircleFill = FALSE, ellipseFill = FALSE,
    lab = NULL
)
bplot
```

```{r pcatools pairs -annotated-, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.width=12}
plot_legend <- get_legend(bplot)
pplot <- pairsplot(pca.metabs2, colby = "fullClass", colkey = color_key$fullClass)
print(cowplot::plot_grid(pplot, plot_legend, nrow = 2, rel_heights = c(1.9, 0.1)))
```

```{r save data -annotated-, echo=FALSE, message=FALSE, warning=FALSE, eval = TRUE}
write.table(metabs.log, file.path(dir_data, "Plasma_untargetedANNOT_metabs_CRMNstandardsnormalized.txt"), sep = "\t", row.names = F)
```


# Cohort - confirm we have the same numbers
```{r metadata diff, echo=FALSE, message=FALSE, warning=FALSE}
metadata <- metabs.log[, 1:21]
metadata$Statins <- ifelse(metadata$Statins == 0, "No", "Yes")
metadata$`Hormone replacement therapy` <- ifelse(metadata$`Hormone replacement therapy` == 0, "No", "Yes")

library(tableone)
## Vector of variables to summarize
vars_test <- c(
    "Gender", "AgeAtVisit", "BMI", "CognitiveSyndrome", "Diagnosis", "MMSE_Score", "Cerebrospinal.fluids_Amyloid.beta.42.pg.ml",
    "Cerebrospinal.fluids_Phospho.Tau.pg.ml", "Cerebrospinal.fluids_Total.Tau.pg.ml", "Genetics_Risk factors_APOE",
    "Statins", "Hormone replacement therapy", "Neuropsychology_ACE_Total score", "CSF_glucose"
)
## Vector of categorical variables that need transformation
catVars <- c("Gender", "CognitiveSyndrome", "Diagnosis")

tab3 <- CreateTableOne(vars = vars_test, strata = "fullClass", data = metadata, factorVars = catVars)

tab_df <- print(tab3, printToggle = FALSE, noSpaces = TRUE)
tab_df <- tab_df[, -ncol(tab_df)]
DT::datatable(tab_df, options = list(
    scrollX = TRUE,
    pageLength = nrow(tab_df), # or -1 to display all rows if you're using lengthMenu
    lengthMenu = list(c(-1, 10, 25, 50), c("All", "10", "25", "50"))
), class = "cell-border stripe")
```

# Session
```{r session info, echo=FALSE, message=FALSE, warning=FALSE}
pander::pander(sessionInfo())
```