---
title: "Lundbeck project - Unargeted CSF CGMS metabolomics - Rigshospitalet - Univariate analysis"
author: "Marc Clos-Garcia"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: united
    self_contained: true
date: "`r Sys.Date()`"
---

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
# Suppress progress bar output globally
knitr::opts_chunk$set(progress = FALSE)

## libraries
library(here)
library(tidyverse)
library(ggrepel)
library(glmnet)
library(broom)
library(pROC)
library(caret)
library(htmlwidgets)
library(DT)
library(knitr)
library(DALEX)
library(shapviz)
library(ggbeeswarm)


## directories
dir_data <- here("data/CSF")
if (!dir.exists(dir_data)) dir.create(dir_data)
dir_res <- here("results/CSF")
if (!dir.exists(dir_res)) dir.create(dir_res)
dir_code <- here("code")
if (!dir.exists(dir_code)) dir.create(dir_code)

## Set cores
max_cores <- if (Sys.getenv("_R_CHECK_LIMIT_CORES_") != "") 2 else parallel::detectCores() - 1
# Setup the parallel backend with the appropriate number of workers
BPPARAM <- BiocParallel::SnowParam(workers = max_cores)

# Define the JS rowCallback function as a string
rowCallback <- "
  function(row, data, index) {
    // Example: color every even row
    if (index % 2 === 0) {
      $(row).css('background-color', '#f2f2f2');
    }
  }
"
```

```{r functions, echo=FALSE, message=FALSE, warning=FALSE}
## T-test analysis
ttest_analysis <- function(metabolites, metadata, contrast) {
    output_dataframe <- data.frame(
        "mean_diff" = c(),
        "fold_change" = c(),
        "p_value" = c()
    )

    metadata.reduced <- metadata %>% dplyr::filter(fullClass %in% contrast)
    metabolites.reduced <- metabolites[row.names(metabolites) %in% metadata.reduced$sample, ]

    if (!identical(as.character(row.names(metabolites.reduced)), as.character(metadata.reduced$sample))) {
        metadata.reduced <- metadata.reduced[match(row.names(metabolites.reduced), metadata.reduced$sample), ]
    } else {}

    for (metabolite in colnames(metabolites.reduced)) {
        tmp_df <- cbind(metabolites.reduced, "class" = metadata.reduced$fullClass)
        tmp_results <- t.test(tmp_df[[metabolite]] ~ tmp_df[["class"]])
        tmp_mean_diff <- 10^tmp_results$estimate[[1]] - 10^tmp_results$estimate[[2]]
        tmp_fold_change <- 10^(tmp_results$estimate[[1]] - tmp_results$estimate[[2]])
        tmp_pval <- tmp_results$p.value
        output_dataframe[metabolite, "mean_diff"] <- tmp_mean_diff
        output_dataframe[metabolite, "fold_change"] <- tmp_fold_change
        output_dataframe[metabolite, "p_value"] <- tmp_pval
    }
    output_dataframe$Metabolite <- rownames(output_dataframe)
    output_dataframe$log2FC <- log2(output_dataframe$fold_change)
    output_dataframe$FDR <- p.adjust(output_dataframe$p_value, method = "fdr")
    output_dataframe <- output_dataframe[, c("Metabolite", "mean_diff", "fold_change", "log2FC", "p_value", "FDR")]
    return(output_dataframe)
}

## GLM models
glm_analysis <- function(metabolites, metadata, contrast) {
    output_dataframe <- data.frame(
        "mean_diff" = c(),
        "fold_change" = c(),
        "p_value" = c()
    )

    metadata.reduced <- metadata %>% dplyr::filter(fullClass %in% contrast)
    metabolites.reduced <- metabolites[row.names(metabolites) %in% metadata.reduced$sample, ]

    if (!identical(as.character(row.names(metabolites.reduced)), as.character(metadata.reduced$sample))) {
        metadata.reduced <- metadata.reduced[match(row.names(metabolites.reduced), metadata.reduced$sample), ]
    } else {}

    for (metabolite in colnames(metabolites.reduced)) {
        tmp_df <- cbind(metabolites.reduced, "class" = metadata.reduced$fullClass,  "Age" = metadata.reduced$AgeAtVisit, 
                        "Gender" = metadata.reduced$Gender, "Statins" = metadata.reduced$Statins)
        tmp_results <- glm(tmp_df[["class"]] ~ 1 + tmp_df[[metabolite]] + tmp_df[["Age"]] + tmp_df[["Gender"]] + tmp_df[["Statins"]], data = tmp_df, binomial(link="logit"))
        tmp_results_s <- summary(tmp_results)
        output_dataframe[metabolite, "estimate"] <- tmp_results$coefficients[2]
        output_dataframe[metabolite, "se"] <- tmp_results_s$coefficients[2, 2]
        output_dataframe[metabolite, "p_value"] <- tmp_results_s$coefficients[2, 4]
    }
    output_dataframe$Metabolite <- rownames(output_dataframe)
    output_dataframe$FDR <- p.adjust(output_dataframe$p_value, method = "fdr")
    output_dataframe$odds_ratio <- exp(output_dataframe$estimate)
    output_dataframe$lower_ci <- exp(output_dataframe$estimate - 1.96 * output_dataframe$se) # 1,96 for 95% CI
    output_dataframe$upper_ci <- exp(output_dataframe$estimate + 1.96 * output_dataframe$se)

    output_dataframe <- output_dataframe[, c(
        "Metabolite", "estimate", "se", "p_value", "FDR", "odds_ratio",
        "lower_ci", "upper_ci"
    )]

    return(output_dataframe)
}


## LASSO models
lasso_analysis <- function(metabolites, metadata, contrast) {
    metadata.reduced <- metadata %>% dplyr::filter(fullClass %in% contrast)
    metabolites.reduced <- metabolites[row.names(metabolites) %in% metadata.reduced$sample, ]

    if (!identical(as.character(row.names(metabolites.reduced)), as.character(metadata.reduced$sample))) {
        metadata.reduced <- metadata.reduced[match(row.names(metabolites.reduced), metadata.reduced$sample), ]
    } else {}

    # ---- STEP 1: Prepare data ----
    # "class" must be 0/1 (numeric or factor)
    # "metabolite_columns" contains all metabolite names
    X <- as.matrix(metabolites.reduced) # predictor matrix
    y <- as.character(metadata.reduced$fullClass) # outcome (binary)
    y <- match(y, unique(y)) - 1 # convert to 0/1 for LASSO modelling

    # ---- STEP 2: LASSO with cross-validation ----
    set.seed(sample(seq(100, 999), 1)) # for reproducibility
    cvfit <- cv.glmnet(X, y, family = "binomial", alpha = 1, type.measure = "deviance")
    lambda_min <- cvfit$lambda.min

    coef_lasso <- coef(cvfit, s = lambda_min)
    selected_vars <- rownames(coef_lasso)[as.numeric(coef_lasso) != 0]
    selected_vars <- selected_vars[selected_vars != "(Intercept)"]

    cat("Selected metabolites:\n")
    print(selected_vars)

    if (length(selected_vars) > 0) {
        # ---- STEP 3: Refit GLM ----
        tmp_df <- cbind(metabolites.reduced, "class" = metadata.reduced$fullClass)
        formula <- as.formula(paste("class ~", paste(selected_vars, collapse = " + ")))
        final_glm <- glm(formula, data = tmp_df, family = binomial)

        results <- tidy(final_glm) %>%
            mutate(
                odds_ratio = exp(estimate),
                conf.low = exp(estimate - 1.96 * std.error),
                conf.high = exp(estimate + 1.96 * std.error)
            )

        # ---- STEP 4: Cross-validated AUC ----
        set.seed(sample(seq(100, 999), 1))
        tmp_df$class <- droplevels(tmp_df$class)
        folds <- createFolds(tmp_df$class, k = 5, returnTrain = TRUE)

        # Precompute test indices and a per-row fold id (for within-test-fold permutation)
        n <- nrow(tmp_df)
        all_idx <- seq_len(n)
        test_idx_list <- lapply(folds, function(tr) setdiff(all_idx, tr))

        fold_id <- integer(n)
        for (j in seq_along(test_idx_list)) {
            fold_id[test_idx_list[[j]]] <- j
        }
        split_idx <- split(all_idx, fold_id) # rows belonging to each test fold

        cv_auc <- function(data, formula, folds) {
            auc_results <- list()
            aucs <- numeric(length(folds))
            for (j in seq_along(folds)) {
                train_idx <- folds[[j]]
                test_idx <- setdiff(seq_len(nrow(data)), train_idx)

                train_data <- data[train_idx, ]
                test_data <- data[test_idx, ]

                fit <- glm(formula, data = train_data, family = binomial)
                preds <- predict(fit, newdata = test_data, type = "response")

                ytest <- droplevels(test_data$class)
                if (length(unique(ytest)) < 2L) {
                    # Degenerate test set — should not happen with within-fold permutation,
                    # but we guard anyway.
                    aucs[j] <- NA_real_
                } else {
                    # You can add `positive = "<case_label>"` if you want to be explicit
                    roc_obj <- pROC::roc(response = ytest, predictor = preds, quiet = TRUE)
                    aucs[j] <- as.numeric(pROC::auc(roc_obj))
                    auc_results[[j]] <- data.frame(sensitivity = roc_obj$sensitivities, specificity = roc_obj$specificities, fdr = 1 - roc_obj$specificities)
                }
            }
            mean_aucs <- mean(aucs, na.rm = TRUE)
            return(list("aucs" = aucs, "mean_auc" = mean_aucs, "aucs_values" = auc_results))
        }

        observed_auc <- cv_auc(tmp_df, formula, folds)
        cat("Observed mean CV AUC:", observed_auc$mean_auc, "\n")

        # ---- STEP 5: Permutation Test ----
        set.seed(sample(seq(100, 999), 1))
        n_perm <- 1000
        perm_aucs <- list()

        for (i in seq_len(n_perm)) {
            df_perm <- tmp_df
            # KEY: permute labels *within each test fold* to preserve per-fold class balance
            for (idx in split_idx) {
                df_perm$class[idx] <- sample(df_perm$class[idx])
            }
            perm_aucs[[i]] <- cv_auc(df_perm, formula, folds)
        }

        # Unbiased permutation p-value
        perm_aucs_mean <- lapply(perm_aucs, function(x) {
            x$mean_auc
        })
        perm_p_value <- (1 + sum(perm_aucs_mean >= observed_auc$mean_auc, na.rm = TRUE)) / (1 + n_perm)
        cat("Permutation p-value:", perm_p_value, "\n")

        # ---- STEP 6: ROC Curve (final model) ----
        full_preds <- predict(final_glm, type = "response")
        # If you want an explicit positive class, set positive = "CaseLabel"
        full_roc <- roc(tmp_df$class, full_preds, quiet = TRUE)
        #  plot(full_roc, main = paste0("ROC Curve (AUC = ", round(auc(full_roc), 3), ")"))

        if (length(selected_vars) > 1) {
            # ---- STEP 6: Shapley scores/plot ----
            # Convert class to 0/1 (numeric)
            y_numeric <- as.numeric(tmp_df$class) - 1
            # Prepare explainer
            explainer <- explain(
                model = final_glm,
                data = tmp_df[, selected_vars],
                y = y_numeric,
                label = "GLM Model"
            )

            # Compute SHAP values for multiple rows
            # This loops over each row and collects SHAP values
            shap_list <- lapply(1:nrow(tmp_df), function(i) {
                parts <- predict_parts(explainer, new_observation = tmp_df[i, selected_vars], type = "shap", B = 10)
                # Extract SHAP values and reorder by variable
                shap_values <- setNames(parts$contribution, parts$variable_name)
                shap_values[selected_vars] # Ensure consistent ordering
            })
            shap_df <- do.call(rbind, shap_list)
            # Combine all into one shapviz object
            sv <- shapviz(shap_df, X = tmp_df[, selected_vars])


            # ---- STEP 7: save output ----
            output <- list(
                "Selected_metabolites" = results,
                "Observed_AUCs" = observed_auc,
                "Permutation_AUCs" = perm_aucs,
                "Permutation_pval" = perm_p_value,
                "Final_model" = full_roc,
                "SHAP_values" = shap_df,
                "SHAP_viz" = sv
            )
        } else {
            # ---- STEP 7: save output ----
            output <- list(
                "Selected_metabolites" = results,
                "Observed_AUCs" = observed_auc,
                "Permutation_AUCs" = perm_aucs,
                "Permutation_pval" = perm_p_value,
                "Final_model" = full_roc,
                "SHAP_values" = NA,
                "SHAP_viz" = NA
            )
        }
    } else {
        cat("No metabolites selected at lambda.min.\n")
        output <- list(
            "Selected_metabolites" = NA,
            "Observed_AUCs" = NA,
            "Permutation_AUCs" = NA,
            "Permutation_pval" = NA,
            "Final_model" = NA,
            "SHAP_values" = NA,
            "SHAP_viz" = NA
        )
    }


    return(output)
}
```


```{r load data, echo=FALSE, message=FALSE, warning=FALSE}
cohort <- read.table(file.path(dir_data, "CSF_untargetedANNOT_metabs_CRMNstandardsnormalized.txt"), sep = "\t", header = T)

## correct fullClass cohort
cohort$fullClass <- factor(cohort$fullClass, levels = c("HC", "MCI_Non-AD", "MCI", "AD", "VaD"))

## split between metadata and metabolomics data
metadata <- cohort[, 1:22]
row.names(metadata) <- cohort$sample
metabolites <- cohort[, 24:ncol(cohort)]
row.names(metabolites) <- cohort$sample


## set up color key
color_key <- list(fullClass = c(
    "HC" = "skyblue2",
    "MCI_Non-AD" = "darkolivegreen3",
    "MCI" = "goldenrod1",
    "AD" = "salmon",
    "VaD" = "darkred"
))

## define contrasts and comparisons
contrasts <- levels(cohort$fullClass)
comparisons <- list(
    "HC" = list(
        c("HC", "MCI_Non-AD"), c("HC", "MCI"), c("HC", "AD"), c("HC", "VaD")
    ),
    "MCI_Non-AD" = list(
        c("MCI_Non-AD", "MCI"), c("MCI_Non-AD", "AD"), c("MCI_Non-AD", "VaD")
    ),
    "MCI" = list(
        c("MCI", "AD"), c("MCI", "VaD")
    ),
    "AD" = list(
        c("AD", "VaD")
    )
)
```


# Context
t-test
linear/logistic models
RandomForest?

# Metabolite ratios

* Brain metabolism shift towards ketone oxidation from glucose: X3.Hydroxy.butyrate/pyruvic acid
* Glycolytic perturbation - downstream oxidation: glyceric acid / pyruvic acid 
* Redox / short hydroxy:keto balance: X2.3.Dihydroxybutanoic acid/X2.Keto.3.methylvaleric acid or X2.3.Dihydroxybutanoic acid/pyruvic acid
* phospholipid / mebrane turnover: O.phosphoethanolamine / phosphoric acid or O.Phosphoethanolamine / X1.5.Anhydrohexitol
* Desaturation / lipid composition index: Palmitoleic.acid.TMS / Tetradecanoic acid or Myristoleic acid / tetradecanoic acid
* Glucose/polyol pathway in CSF - polyol pathway contributed to osmotic/oxidative stress in diabetes, a risk factor for dementia: d.Glucose / Sorbitol or d.Glucose/meso.Erythitol
* Tryptophan / kynurenine-type (tryp status): L.Tryptophan / ribonic.acid or X3.Deoxy.erythro.pentitol
* Nitrogen clearance: creatinine/urea
* Purine oxidative stress / antioxidant: Uric.acid/Urea or Uric.acid/Creatinine
* Branched-Chain Amino Acid and Ketone catabolisme:
  ** L.Norvaline / L.Norleucine
* TCA and anaplerosis: 
  ** glyceric acid/ ribonic acid
  ** pyruvic acid/malic.acid.1.ethyl.ester
* Osmoregulation/nitrogen handling: Ornithine/Urea
* Lipid and membrane integrity:
  ** dodecanoic acid / heptadecanoic acid
  ** palmitoleic acid TMS / pentadecanoic acid
* Neurotransmitters and polyol pathways>
  ** ethanolamine / O.Phosphoethanolamine
  ** X2.Aminoethanol / Ethanolamine
  ** d.Galactose / D.Ribulose
* Oxidative and aromatic pathways:
  ** Benzoic.acid / Quinic.acid
  ** Quinic.acid / Uric.acid

```{r}
metabolites <- cbind(metabolites,
                     hydroxybut.pyruvic = metabolites$X3.Hydroxy.butyrate - metabolites$Pyruvic.acid,
                     glyceric.pyruvic = metabolites$Glyceric.acid - metabolites$Pyruvic.acid, 
                     hydroxybutanoic.keto = metabolites$X2.3.Dihydroxybutanoic.acid - metabolites$X2.Keto.3.methylvaleric.acid, 
                     hydroxybutanoic.pyruvic = metabolites$X2.3.Dihydroxybutanoic.acid - metabolites$Pyruvic.acid,
                     ph_ethanolamine.phosphoric = metabolites$O.Phosphoethanolamine - metabolites$Phosphoric.acid,
                     ph_ethanolamine.anhydroexitol = metabolites$O.Phosphoethanolamine - metabolites$X1.5.Anhydrohexitol,
                     TMS.tetradecanoic = metabolites$Palmitoleic.acid.TMS - metabolites$Tetradecanoic.acid,
                     myristoleic.tetradecanoic = metabolites$Myristoleic.acid - metabolites$Tetradecanoic.acid,
                     gluc.sorbitol = metabolites$d.Glucose - metabolites$Sorbitol,
                     gluc.erythitol = metabolites$d.Glucose - metabolites$meso.Erythritol,
                     tryp.ribonic = metabolites$L.Tryptophan - metabolites$Ribonic.acid,
                     tryp.pentitol = metabolites$L.Tryptophan - metabolites$X3.Deoxy.erythro.pentitol,
                     creat.urea = metabolites$Creatinine - metabolites$Urea,
                     uric_acid.urea = metabolites$Uric.acid - metabolites$Urea,
                     uric_acid.creat = metabolites$Uric.acid - metabolites$Creatinine,
                     norvaline.norleucine = metabolites$L.Norvaline - metabolites$L.Norleucine,
                     glyceric.ribonic = metabolites$Glyceric.acid - metabolites$Ribonic.acid,
                     pyruvic.malic_acid = metabolites$Pyruvic.acid - metabolites$Malic.acid.1.ethyl.ester,
                     ornithine.urea = metabolites$Ornithine - metabolites$Urea,
                     dodecanoic.heptadecanoic = metabolites$Dodecanoic.acid - metabolites$Heptadecanoic.acid,
                     TMS.pentadecanoic = metabolites$Palmitoleic.acid.TMS - metabolites$Pentadecanoic.acid,
                     ethanolamine.ph_ethanolamine = metabolites$Ethanolamine - metabolites$O.Phosphoethanolamine,
                     aminoeth.ethanolamine = metabolites$X2.Aminoethanol - metabolites$Ethanolamine,
                     galactose.ribulose = metabolites$d.Galactose - metabolites$D.Ribulose,
                     benzoic.quinic = metabolites$Benzoic.acid - metabolites$Quinic.acid,
                     quinic.uric = metabolites$Quinic.acid - metabolites$Uric.acid)


row.names(metabolites) == row.names(metadata)
cohort2 <- cbind(metadata, metabolites) 
write.table(cohort2, file.path(dir_data, "CSF_CRMNnormalized_ratios_UNtargeted.txt"), sep = "\t", row.names = F)
```

```{r stupidly needed to knit tables, echo=FALSE,include = FALSE}
# You need this code to conduct the magic dependences attaching...
DT::datatable(matrix())
```


# T-test analysis {.tabset}
```{r ttest run, echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
ttest_results <- lapply(comparisons, function(contrast) {
    results <- lapply(contrast, function(comparison) {
        res <- ttest_analysis(metabolites, metadata, comparison)
        return(res)
    })
    names(results) <- unlist(lapply(contrast, function(x) {
        paste(x, collapse = "vs")
    }))
    combined_results <- do.call(rbind, Map(function(df, name) {
        df$source <- name
        df
    }, results, names(results)))

    return(combined_results)
})

ttest_results <- do.call(rbind, ttest_results)
ttest_results <- ttest_results |> dplyr::select(source, Metabolite, everything())

for (comp in names(comparisons)) {
    comp.list <- comparisons[[comp]]
    for (subc in comp.list) {
        cat("## ", paste(subc, collapse = "vs"), "\n\n")
        ttest_specific <- ttest_results |> dplyr::filter(source == paste(subc, collapse = "vs"))
        raw.p <- ggplot(ttest_specific, aes(log2FC, -log10(p_value))) +
            geom_point() +
            geom_text_repel(
                data = subset(ttest_specific, p_value <= 0.05),
                aes(label = Metabolite)
            ) +
            geom_hline(yintercept = -log10(0.05)) +
            theme_bw() +
            labs(
                x = bquote(~ log[2] ~ "Fold-Change"),
                y = bquote("-" ~ log[10] ~ "pvalue")
            ) +
            labs(title = paste(subc, collapse = "vs"))
        fdr.p <- ggplot(ttest_specific, aes(log2FC, -log10(FDR))) +
            geom_point() +
            geom_text_repel(
                data = subset(ttest_specific, FDR <= 0.05),
                aes(label = Metabolite)
            ) +
            geom_hline(yintercept = -log10(0.05)) +
            theme_bw() +
            labs(
                x = bquote(~ log[2] ~ "Fold-Change"),
                y = bquote("-" ~ log[10] ~ "FDR")
            ) +
            labs(title = paste(subc, collapse = "vs"))
        print(ggpubr::ggarrange(raw.p, fdr.p))
        cat("\n\n")
        ttest_specific <- ttest_specific |>
            arrange(p_value) |>
            mutate_if(is.numeric, round, 3)
        cat(knitr::knit_print(DT::datatable(ttest_specific,
            escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
            options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
        )))
    }
}
```

# Logistic regression models {.tabset}
```{r logistic models, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
glm_results <- lapply(comparisons, function(contrast) {
    results <- lapply(contrast, function(comparison) {
        res <- glm_analysis(metabolites, metadata, comparison)
        return(res)
    })
    names(results) <- unlist(lapply(contrast, function(x) {
        paste(x, collapse = "vs")
    }))
    combined_results <- do.call(rbind, Map(function(df, name) {
        df$source <- name
        df
    }, results, names(results)))

    return(combined_results)
})
```


```{r glm show, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results = "asis", fig.width=6, fig.height=4}
## show results
for (n in names(glm_results)) {
    cat("## ", n, " {.tabset} \n\n")
    df <- glm_results[[n]]
    df_list <- list()
    for (contrast in unique(df$source)) {
        cat("### ", contrast, "\n\n")
        df_red <- df[df$source == contrast, ]
        df_red$source <- NULL
        df_red <- df_red |>
            dplyr::mutate_if(is.numeric, round, 2) |>
            as.data.frame()
        cat(knitr::knit_print(DT::datatable(df_red,
            escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
            options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
        )))
        cat("\n\n")
        df_red <- df_red[df_red$p_value <= 0.05, ]
        if (nrow(df_red) > 0) {
            forest_plot <- df_red |>
                dplyr::mutate(significance = case_when(
                    p_value <= 0.05 ~ "*",
                    p_value <= 0.01 ~ "**",
                    p_value <= 0.001 ~ "***",
                    TRUE ~ ""
                )) |>
                ggplot(aes(odds_ratio, Metabolite)) +
                geom_point() +
                geom_errorbar(aes(xmin = lower_ci, xmax = upper_ci), width = 0.15) +
                geom_vline(xintercept = 1, linetype = "dashed") +
                geom_text(aes(x = max(upper_ci, na.rm = T) + 0.25, label = significance),
                    vjust = .75,
                    size = 8, # Size of stars
                    na.rm = TRUE
                ) +
                ylab("Odds Ratio (95% CI)") +
                xlab("") +
                theme_minimal() +
                theme(
                    axis.line = element_line(),
                    strip.background = element_rect(),
                    panel.background = element_rect()
                )
            print(forest_plot)
            cat("\n\n")
        } else {
            NULL
        }
    }
}
```



# LASSO regression
We noticed that the SE of the metabolites in the logistic models are disproportionaly high, which is a sign of low signal or not enough power in the analyzed dataset.
We decided to try with more complex models, i.e. LASSO regression approaches, that will combine all relevant metabolites in a single model. It will also allow us to identify the higher contributing metabolites.
```{r, echo = FALSE, message=FALSE, warning=FALSE}
lasso_results <- lapply(comparisons, function(contrast) {
    results <- lapply(contrast, function(comparison) {
        res <- lasso_analysis(metabolites, metadata, comparison)
        return(res)
    })
    names(results) <- unlist(lapply(contrast, function(x) {
        paste(x, collapse = "vs")
    }))
    combined_results <- do.call(rbind, Map(function(df, name) {
        df$source <- name
        df
    }, results, names(results)))

    return(combined_results)
})
```

## AUCs {.tabset}
```{r AUC boxplots, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
for (n in names(lasso_results)) {
    cat("### ", n, " {.tabset} \n\n")
    lasso.df <- lasso_results[[n]]
    for (contrast in rownames(lasso.df)) {
        cat("#### ", contrast, "\n\n")
        lasso.df2 <- lasso.df[contrast, ]
        if (is.null(ncol(lasso.df2$Selected_metabolites))) {
            cat("No selected metabolites in the minimal lambda")
            cat("\n\n")
        } else {
            obs_aucs <- lasso.df2$Observed_AUCs$aucs
            permutation_aucs <- unlist(lapply(lasso.df2$Permutation_AUCs, function(au) {
                return(au$aucs)
            }))
            final_auc <- lasso.df2$Final_model$auc
            AUCs_plot <- rbind(
                data.frame(AUC = obs_aucs, source = "Observed_AUC"),
                data.frame(AUC = permutation_aucs, source = "Permutation_AUC"),
                data.frame(AUC = final_auc, source = "Final_model_AUC")
            )
            AUCs_plot$source <- factor(AUCs_plot$source, levels = c("Final_model_AUC", "Permutation_AUC", "Observed_AUC"))
            auc_bplot <- ggplot(AUCs_plot, aes(AUC, source)) +
                ggbeeswarm::geom_quasirandom() +
                geom_boxplot(alpha = .25) +
                geom_vline(xintercept = 0.5, color = "darkred") +
                theme_bw() +
                xlim(0, 1) +
                labs(y = "AUC origin", title = contrast)
            print(auc_bplot)
            cat("\n\n")
            sel_mets <- lasso.df2$Selected_metabolites |>
                dplyr::mutate_if(is.numeric, round, 3) |>
                as.data.frame()

            cat(knitr::knit_print(DT::datatable(sel_mets,
                escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
                options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
            )))
            cat("\n\n")

            shap_plot <- shapviz::sv_importance(lasso.df2$SHAP_viz, kind = "beeswarm") + theme_bw()
            print(shap_plot)
            cat("\n\n")
        }
    }
}
```

# Regression against continous variables {.tabset}
```{r  echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
continuous_analysis <- function(metabolites, metadata, variable) {
    output_dataframe <- data.frame()

    metadata.reduced <- metadata %>% dplyr::filter(!is.na(metadata[[variable]]))
    metadata.reduced <- metadata.reduced %>% dplyr::filter(metadata.reduced[[variable]] > 1)
    metabolites.reduced <- metabolites[row.names(metabolites) %in% metadata.reduced$sample, ]

    if (!identical(as.character(row.names(metabolites.reduced)), as.character(metadata.reduced$sample))) {
        metadata.reduced <- metadata.reduced[match(row.names(metabolites.reduced), metadata.reduced$sample), ]
    } else {}

    for (metabolite in colnames(metabolites.reduced)) {
        # numeric_var <- log(metadata.reduced[[variable]])
        # numeric_var <- scale(numeric_var)
        numeric_var <- scale(metadata.reduced[[variable]])
        tmp_df <- cbind(metabolites.reduced, "numeric_var" = numeric_var, Age = metadata.reduced[["AgeAtVisit"]], Gender = metadata.reduced[["Gender"]], statins = metadata.reduced[["Statins"]])
        tmp_df[[metabolite]] <- scale(tmp_df[[metabolite]], center = TRUE, scale = TRUE)
        tmp_results <- glm(tmp_df[["numeric_var"]] ~ 1 + tmp_df[[metabolite]] + tmp_df[["Age"]] + tmp_df[["Gender"]] + tmp_df[["statins"]], data = tmp_df, family = "gaussian")
        tmp_results_s <- summary(tmp_results)
        null_deviance <- tmp_results$null.deviance
        residual_deviance <- tmp_results$deviance
        # Calculate R-squared
        r_squared <- 1 - (residual_deviance / null_deviance)
        CIs <- confint(tmp_results)

        output_dataframe[metabolite, "r2"] <- r_squared
        output_dataframe[metabolite, "estimate"] <- tmp_results$coefficients[2]
        output_dataframe[metabolite, "se"] <- tmp_results_s$coefficients[2, 2]
        output_dataframe[metabolite, "CI_upper"] <- CIs[2, 2]
        output_dataframe[metabolite, "CI_lower"] <- CIs[2, 1]
        output_dataframe[metabolite, "p_value"] <- tmp_results_s$coefficients[2, 4]
    }
    output_dataframe$Metabolite <- rownames(output_dataframe)
    output_dataframe$FDR <- p.adjust(output_dataframe$p_value, method = "fdr")

    output_dataframe <- output_dataframe[, c(
        "Metabolite", "r2", "estimate", "se", "CI_upper", "CI_lower", "p_value", "FDR"
    )]

    return(output_dataframe)
}
```



```{r  echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
metadata <- metadata |>
    dplyr::rename(
        "betaAmyloid" = "Cerebrospinal.fluids_Amyloid.beta.42.pg.ml",
        "phosphoTau" = "Cerebrospinal.fluids_Phospho.Tau.pg.ml",
        "totalTau" = "Cerebrospinal.fluids_Total.Tau.pg.ml",
        ACE_Score = Neuropsychology_ACE_Total.score
    )
variables <- c("MMSE_Score", "betaAmyloid", "phosphoTau", "totalTau", "ACE_Score", "BMI", "CSF_glucose")
```


```{r linear models, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
linear_results <- lapply(variables, function(comparison) {
    res <- continuous_analysis(metabolites, metadata, comparison)
    return(res)
})

names(linear_results) <- variables
```


```{r linear show, echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
## show results
for (n in names(linear_results)) {
    cat("## ", n, " {.tabset} \n\n")
    df <- linear_results[[n]]
    df <- df |>
        dplyr::mutate_if(is.numeric, round, 2) |>
        as.data.frame()
    cat(knitr::knit_print(DT::datatable(df,
        escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
    cat("\n\n")
    df_red <- df[df$p_value <= 0.05, ]
    metabolites_plot <- cbind(metadata, metabolites)
    metabolites_plot <- metabolites_plot %>% dplyr::filter(!is.na(metabolites_plot[[n]]))
    metabolites_plot <- metabolites_plot %>% dplyr::filter(metabolites_plot[[n]] > 1)
    metabolites_plot[[n]] <- log(metabolites_plot[[n]])
    for (met in unique(df_red$Metabolite)) {
        cat("### ", met, "\n\n")
        p <- ggplot(metabolites_plot, aes_string(x = n, y = met)) +
            geom_point() +
            geom_smooth(method = "lm", se = F) +
            theme_bw() +
            labs(
                x = paste0("log-transformed ", n),
                y = paste0(met, " normalized abundance")
            )
        print(p)
        cat("\n\n")
    }
}
```


# Regression against continous variables - HC+MCI+AD {.tabset}

```{r linear models 2, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
metadata.red <- metadata %>% dplyr::filter(fullClass %in% c("HC", "MCI", "AD"))
metabolites.red <- metabolites[metadata.red$sample, ]
linear_results <- lapply(variables, function(comparison) {
    res <- continuous_analysis(metabolites, metadata, comparison)
    return(res)
})

names(linear_results) <- variables
```

```{r stupidly needed to knit tables3b, echo=FALSE,include = FALSE}
# You need this code to conduct the magic dependences attaching...
DT::datatable(matrix())
```

```{r linear show2, echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
## show results
for (n in names(linear_results)) {
    cat("## ", n, " {.tabset} \n\n")
    df <- linear_results[[n]]
    df <- df |>
        dplyr::mutate_if(is.numeric, round, 2) |>
        as.data.frame()
    cat(knitr::knit_print(DT::datatable(df,
        escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
    cat("\n\n")
    df_red <- df[df$p_value <= 0.05, ]
    metabolites_plot <- cbind(metadata, metabolites)
    metabolites_plot <- metabolites_plot %>% dplyr::filter(!is.na(metabolites_plot[[n]]))
    metabolites_plot <- metabolites_plot %>% dplyr::filter(metabolites_plot[[n]] > 1)
    metabolites_plot[[n]] <- log(metabolites_plot[[n]])
    for (met in unique(df_red$Metabolite)) {
        cat("### ", met, "\n\n")
        p <- ggplot(metabolites_plot, aes_string(x = n, y = met)) +
            geom_point() +
            geom_smooth(method = "lm", se = F) +
            theme_bw() +
            labs(
                x = paste0("log-transformed ", n),
                y = paste0(met, " normalized abundance")
            )
        print(p)
        cat("\n\n")
    }
}
```

# Regression against continous variables - MCI {.tabset}
```{r linear models 3, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
metadata.red <- metadata %>% dplyr::filter(fullClass %in% c("MCI"))
metabolites.red <- metabolites[metadata.red$sample, ]
linear_results <- lapply(variables, function(comparison) {
    res <- continuous_analysis(metabolites, metadata, comparison)
    return(res)
})

names(linear_results) <- variables
```

```{r stupidly needed to knit tables3c, echo=FALSE,include = FALSE}
# You need this code to conduct the magic dependences attaching...
DT::datatable(matrix())
```

```{r linear show3, echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
## show results
for (n in names(linear_results)) {
    cat("## ", n, " {.tabset} \n\n")
    df <- linear_results[[n]]
    df <- df |>
        dplyr::mutate_if(is.numeric, round, 2) |>
        as.data.frame()
    cat(knitr::knit_print(DT::datatable(df,
        escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
    cat("\n\n")
    df_red <- df[df$p_value <= 0.05, ]
    metabolites_plot <- cbind(metadata, metabolites)
    metabolites_plot <- metabolites_plot %>% dplyr::filter(!is.na(metabolites_plot[[n]]))
    metabolites_plot <- metabolites_plot %>% dplyr::filter(metabolites_plot[[n]] > 1)
    metabolites_plot[[n]] <- log(metabolites_plot[[n]])
    for (met in unique(df_red$Metabolite)) {
        cat("### ", met, "\n\n")
        p <- ggplot(metabolites_plot, aes_string(x = n, y = met)) +
            geom_point() +
            geom_smooth(method = "lm", se = F) +
            theme_bw() +
            labs(
                x = paste0("log-transformed ", n),
                y = paste0(met, " normalized abundance")
            )
        print(p)
        cat("\n\n")
    }
}
```

# Regression against continous variables - HC+VaD {.tabset}
```{r linear models 4, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
metadata.red <- metadata %>% dplyr::filter(fullClass %in% c("HC", "VaD"))
metabolites.red <- metabolites[metadata.red$sample, ]
linear_results <- lapply(variables, function(comparison) {
    res <- continuous_analysis(metabolites, metadata, comparison)
    return(res)
})

names(linear_results) <- variables
```

```{r stupidly needed to knit tables3d, echo=FALSE,include = FALSE}
# You need this code to conduct the magic dependences attaching...
DT::datatable(matrix())
```

```{r linear show4, echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
## show results
for (n in names(linear_results)) {
    cat("## ", n, " {.tabset} \n\n")
    df <- linear_results[[n]]
    df <- df |>
        dplyr::mutate_if(is.numeric, round, 2) |>
        as.data.frame()
    cat(knitr::knit_print(DT::datatable(df,
        escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
    cat("\n\n")
    df_red <- df[df$p_value <= 0.05, ]
    metabolites_plot <- cbind(metadata, metabolites)
    metabolites_plot <- metabolites_plot %>% dplyr::filter(!is.na(metabolites_plot[[n]]))
    metabolites_plot <- metabolites_plot %>% dplyr::filter(metabolites_plot[[n]] > 1)
    metabolites_plot[[n]] <- log(metabolites_plot[[n]])
    for (met in unique(df_red$Metabolite)) {
        cat("### ", met, "\n\n")
        p <- ggplot(metabolites_plot, aes_string(x = n, y = met)) +
            geom_point() +
            geom_smooth(method = "lm", se = F) +
            theme_bw() +
            labs(
                x = paste0("log-transformed ", n),
                y = paste0(met, " normalized abundance")
            )
        print(p)
        cat("\n\n")
    }
}
```

# Regression against continous variables - HC+MCI+AD+VaD {.tabset}

```{r linear model relevantsampless, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
metadata.red <- metadata %>% dplyr::filter(fullClass %in% c("HC", "MCI", "AD", "VaD"))
metabolites.red <- metabolites[metadata.red$sample, ]
linear_results <- lapply(variables, function(comparison) {
    res <- continuous_analysis(metabolites, metadata, comparison)
    return(res)
})


names(linear_results) <- variables
```

```{r stupidly needed to knit tablesfinal, echo=FALSE,include = FALSE}
# You need this code to conduct the magic dependences attaching...
DT::datatable(matrix())
```

```{r linear showntsamples releva, echo=FALSE, message=FALSE, warning=FALSE, results = "asis"}
## show results
for (n in names(linear_results)) {
    cat("## ", n, " {.tabset} \n\n")
    df <- linear_results[[n]]
    df <- df |>
        dplyr::mutate_if(is.numeric, round, 2) |>
        as.data.frame()
    cat(knitr::knit_print(DT::datatable(df,
        escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
    cat("\n\n")
    df_red <- df[df$p_value <= 0.05, ]
    metabolites_plot <- cbind(metadata, metabolites)
    metabolites_plot <- metabolites_plot %>% dplyr::filter(!is.na(metabolites_plot[[n]]))
    metabolites_plot <- metabolites_plot %>% dplyr::filter(metabolites_plot[[n]] > 1)
    metabolites_plot[[n]] <- log(metabolites_plot[[n]])
    for (met in unique(df_red$Metabolite)) {
        cat("### ", met, "\n\n")
        p <- ggplot(metabolites_plot, aes_string(x = n, y = met)) +
            geom_point() +
            geom_smooth(method = "lm", se = F) +
            theme_bw() +
            labs(
                x = paste0("log-transformed ", n),
                y = paste0(met, " normalized abundance")
            )
        print(p)
        cat("\n\n")
    }
}
```






# Bayesian approach
```{r, message=FALSE, warning=FALSE, echo=FALSE, results="asis"}
bayesian_results <- list.files(dir_res, pattern = "bayesian")

for(b in bayesian_results){
  bayesian <- read.table(file.path(dir_res, b), header = T, sep = ",")
  
  ## prep title
  comparison <- gsub("CSF_(.*)_bayesian_results\\.csv", "\\1", b)
  comparison <- gsub("_", "vs", comparison)

  cat("## ", comparison, "\n\n")
  bayesian$pval <- 1 - pmax(bayesian$p_pos, bayesian$p_neg)
  bayesian$pval <- ifelse(bayesian$pval == 0, 0.000001, bayesian$pval)
  volcano <- ggplot(bayesian, aes(x = mean_beta, y = -log10(pval), color = significant)) +
    geom_point() +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
      title = paste0("Parallel Bayesian Differential Metabolite Analysis - ", comparison),
      x = "Posterior Mean Effect (β)",
      y = "-log10(1 - Posterior Probability)"
    ) +
    theme_bw() +
    geom_text_repel(
      data = subset(bayesian, (1 - pmax(p_pos, p_neg)) <= 0.05),
      aes(label = Metabolite)
    ) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed")
  print(volcano)
  cat("\n\n")
  cat(knitr::knit_print(DT::datatable(bayesian,
        escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))
  cat("\n\n")
}


all_bayesian <- lapply(bayesian_results, function(b){
  df <- read.table(file.path(dir_res, b), header = T, sep = ",")
  return(df)
})

names(all_bayesian) <- sapply(bayesian_results, function(b){
  comparison <- gsub("CSF_(.*)_bayesian_results\\.csv", "\\1", b)
  comparison <- gsub("_", "vs", comparison)
})

all_bayesian <- do.call(rbind, all_bayesian) %>% 
  tibble::rownames_to_column("source") %>% 
  tidyr::separate(source, sep = "\\.", into = "source") %>% 
  as.data.frame()

```


# Results coherence {.tabset}
```{r, echo=FALSE, warning=FALSE, message=FALSE, results="asis"}
library(UpSetR)
library(ComplexHeatmap)

comparisons <- list("HC_MCI" = "HCvsMCI",
                    "HC_AD" = "HCvsAD", 
                    "HC_VaD" = "HCvsVaD")

diff_mets_list <- list()
for(comp in comparisons){
  cat("##", comp, "\n\n")
  
  bayesian <- all_bayesian %>% dplyr::filter(source == comp) 
  bayesian$pval <- 1 - pmax(bayesian$p_pos, bayesian$p_neg)
  bayesian$pval <- ifelse(bayesian$pval == 0, 0.000001, bayesian$pval)
  
  
  diff.exp.mets <- list(
    "ttest" = ttest_results %>% dplyr::filter(source == comp) %>% dplyr::filter(p_value <= 0.05) %>% pull(Metabolite),
    "GLM" = glm_results$HC %>% dplyr::filter(source == comp) %>% dplyr::filter(p_value <= 0.05) %>%  pull(Metabolite),
    "Lasso" = as.data.frame(lasso_results$HC[comp, "Selected_metabolites"])$term[-1],
    "Bayesian" = bayesian %>% dplyr::filter(pval <= 0.05) %>% pull(Metabolite),
    "GLM_betaAmyloid" = linear_results$betaAmyloid %>% dplyr::filter(p_value <= 0.05) %>% pull(Metabolite),
    "GLM_pTau" = linear_results$phosphoTau %>% dplyr::filter(p_value <= 0.05) %>% pull(Metabolite),
    "GLM_tTau" = linear_results$totalTau %>% dplyr::filter(p_value <= 0.05) %>% pull(Metabolite)
  )
  
  up_plot <- upset(fromList(diff.exp.mets), nintersects = NA, nsets = length(names(diff.exp.mets)))
  print(up_plot)
  cat("\n\n")
  
  diff.exp.mets.m <- ComplexHeatmap::list_to_matrix(diff.exp.mets) %>%  as.data.frame()
  diff.exp.mets.m[] <- as.data.frame(sapply(diff.exp.mets.m, as.integer))  
  diff.exp.mets.m <- diff.exp.mets.m %>% 
    dplyr::mutate(TOTAL = rowSums(.)) %>% 
    dplyr::arrange(-TOTAL) %>% 
    tibble::rownames_to_column("Metabolite") 
  
  diff_mets_list[[comp]] <- diff.exp.mets.m
  
  cat(knitr::knit_print(DT::datatable(diff.exp.mets.m,
        escape = T, height = "100%", class = "cell-border stripe", rownames = FALSE,
        options = list(dom = "Blfrtip", scrollX = TRUE, .noWS = NULL)
    )))

  cat("\n\n")
}

```


# Boxplots of identified differential metabolites {.tabset}
```{r echo=FALSE, results='asis'}
library(ggpubr)


for(comp in comparisons){
  cat("##", comp, " {.tabset} \n\n")
  
  mets.plot <- diff_mets_list[[comp]] %>% 
    dplyr::filter(TOTAL > 1) %>% 
    pull(Metabolite)
  
  comp_plot <- unlist(strsplit(comp, split = "vs"))

  for(m in mets.plot){
    df <- cbind(metadata, metabolite = metabolites[[m]])
    df <- df[df$fullClass %in% comp_plot, ]
    bplots <- ggplot(df, aes(fullClass, metabolite)) +
      geom_boxplot() +
      geom_quasirandom(alpha = .25) +
      stat_compare_means() +
      labs(x = "Clinical group",
           y = paste0("Normalized ", m, " abundance"),
           title = comp) +
      theme_bw()
    
    cat("### ", m, "\n\n")
    print(bplots)
    cat("\n\n")
  }
  
  bayesian <- all_bayesian %>%  dplyr::filter(source == comp) 
  bayesian$pval <- 1 - pmax(bayesian$p_pos, bayesian$p_neg)
  bayesian$pval <- ifelse(bayesian$pval == 0, 0.000001, bayesian$pval)
  
  mets.plot <- bayesian %>%
    dplyr::filter(source == comp) %>% 
    dplyr::filter(Metabolite %in% mets.plot) %>% 
    dplyr::select(Metabolite, mean_beta, pval)


  write.table(mets.plot, file.path(dir_res, paste0("CSF_", comp, "_relevanthits.txt")), sep = "\t", row.names = F)
}
```

# {.unnamed .unlisted}
```{r, echo=FALSE}
pander::pander(sessionInfo())
```